### 2014-07-15 Algorithms/Givens/Givens.tex
updating givens lab

\lab{Algorithms}{Givens rotations and Least squares}{Givens rotations and Least squares}
\objective{Learn how to use orthogonal transformations to perform QR decomposition.}
\label{lab:givens}

In a previous lab, we discussed how to form the QR decomposition of a matrix using a series of householder reflections. The general approach suggested was to apply a series of orthogonal transformations to a given matrix in order to transform it into a matrix of upper triangular form. This same approach can be applied using rotations instead of reflections.

The matrix

\centerline{$\begin{pmatrix}\ cos \theta & - \sin \theta \\ \sin \theta & \cos \theta \end{pmatrix}$}

rotates a vector counterclockwise by $\theta$.
We can rotate a given vector $x = \begin{pmatrix} a \\ b \end{pmatrix}$ into the span of $e_1$ by choosing the correct $\theta$. This zeros out the entry containing $b$.

We could do this by finding the angle between the vector x
%$\begin{pmatrix} a \\ b \end{pmatrix}$ 
and the axis containing $b$ and then rotating the vector by the negative of that angle. However, we want to avoid computations involving $\sin$ and $\arcsin$ at each iteration.  To do so we can solve for $\sin \theta $ and $\cos \theta$ using the pythagorean theorem instead of explicitly computing $\theta$. Such an approach gives us the result that

\centerline{$\cos \theta = \frac{a}{\sqrt{a^2 + b^2}}$ and $\sin \theta = \frac{- b}{\sqrt{a^2 + b^2}}$.}

The transformed image of the vector x is then $\begin{pmatrix} \sqrt{a^2 + b^2} \\ 0 \end{pmatrix}$.

The application of such a rotation to any two rows of a matrix $A$ is an orthonormal transform.
Such a transformation can be represented in matrix form, where $c = \cos \theta$ and $s = \sin \theta$, like this:
\begin{equation*}
\begin{pmatrix}
I & 0 & 0 & 0 & 0 \\
0 & c & 0 & -s & 0 \\
0 & 0 & I & 0 & 0 \\
0 & s & 0 & c & 0 \\
0 & 0 & 0 & 0 & I
\end{pmatrix}
\end{equation*}
In actual computation, we \emph{do not} actually form such a large matrix at each step. Instead, we apply the transformation to the two rows we want to change and then leave the rest of the matrix unchanged.
Proceeding in this manner we can zero out each entry of any matrix below its main diagonal, resulting in a QR factorization.

For example, consider some matrix $A$.
Let $G \left( i, j, k \right)$ denote the Givens rotation that will zero out $A \left[ j, k \right]$ by rotation with row $i$ of $A$.
We can zero out the entries below the main diagonal of A as follows:

\[
\begin{array}{ccccccc}
\begin{pmatrix}
*&*&*\\
*&*&*\\
*&*&*
\end{pmatrix}
&
\underrightarrow{G(1,2,0)}
&\begin{pmatrix}
*&*&*\\
*&*&*\\
0&*&*
\end{pmatrix}
&
\underrightarrow{G(0,1,0)}
&\begin{pmatrix}
*&*&*\\
0&*&*\\
0&*&*
\end{pmatrix}
&
\underrightarrow{G(1,2,1)}
&\begin{pmatrix}
*&*&*\\
0&*&*\\
0&0&*
\end{pmatrix}
\end{array}
\]

Here is an outline for the algorithm.
Let $A$ be the array passed to the function.

\begin{comment}

\begin{algorithm}
\begin{algorithmic}[1]
\Procedure{Givens Triangularization}{$A$}
\State $R \gets \text{copy}(A)$
\State $Q \gets I_A$
\State $G \gets \text(empty)(2,2)$
\For{each column}
    \For{each row below the main diagonal}
        \If{leading element is not zero}
            \State Compute $c$ and $s$ using the entry in the current row and column and the entry immediately above it
            \State Use $c$ and $s$ to construct the matrix $G$
            \State Get a slice of $R$ of the current row and the row above it that includes the columns from the current column onward.
                Multiply it in place by $G$ to zero out the leading nonzero entry of the current row.
            \State Get a slice of the current row of $Q$ and the row above it and apply $G$ to it. (Strictly speaking, you do not need to operate over these entire rows, but the slicing needed to avoid the extra computation is a little more involved, so we will not include that here.)
        \EndIf
    \EndFor
\EndFor
\State \pseudoli{return} $Q^T, R$
\EndProcedure
\end{algorithmic}
\end{algorithm}

\end{comment}

\begin{itemize}[$\bullet$]

\item Make $R$ a copy of $A$ and $Q$ an identity array of the appropriate size.

\item Make an empty $2 \times 2$ array $G$ that will be used to apply the Givens rotations.

\item For each column:

  \begin{itemize}[$\bullet$]

  \item For each row below the main diagonal (starting at the bottom of the column):

    \begin{itemize}[$\bullet$]

    \item If the leading entry of this row is not zero (i.e. if its absolute value is within a given tolerance):

      \begin{itemize}[$\bullet$]

      \item Compute $c$ and $s$ using the entry in the current row and column and the entry immediately above it.

      \item Use $c$ and $s$ to construct the matrix $G$.

      \item Get a slice of $R$ of the current row and the row above it that includes the columns from the current column onward.
      Multiply it in place by $G$ to zero out the leading nonzero entry of the current row.

      \item Get a slice of the current row of $Q$ and the row above it and apply $G$ to it. (Strictly speaking, you do not need to operate over these entire rows, but the slicing needed to avoid the extra computation is a little more involved, so we will not include that here.)

      \end{itemize}

    \end{itemize}

  \end{itemize}

\item Return $Q^T$ and $R$.

\end{itemize}

This version of the algorithm is not the most well-optimized way to use Givens rotations, but it is good for illustrative purposes.

Because Givens rotations only operate on two specific rows at a time, they provide multiple ways to iterate over the array we are processing.
For example, using Givens rotations you could, starting at the bottom of each column and continuing up until you reach the main diagonal, zero out each entry by applying a Givens rotation to each row and the row immediately above it.
Alternatively, within each column, you could could zero out the first entry of each row below the main diagonal by rotating it with the row corresponding to the main diagonal.
Yet, another way to do it would be to start at the bottom left corner of the matrix and then zero out each entry from left to right along each diagonal.
This flexibility is what makes the use of Givens rotations ideal for some problems.
When working with sparse matrices, Givens rotations allow for the solution of linear systems through a series of orthonormal operations that operate only on small parts of the array.
The fact that Givens rotations operate only on small portions of the array also lends well to parallel solutions to systems of equations (or least squares problems).
They allow for the sort of flexibility we had when using Gaussian elimination while still maintaining the favorable stability that comes with using orthonormal transformations.

While Givens rotations do allow for greater flexibility, they also require a greater number of floating point operations than Householder reflections.
In general, the operation count for computing the QR factorization for an $m \times n$ matrix is $3 n^2 \left( m - \frac{n}{3} \right)$.
% Accuracy and Stability of Numerical Algorithms, Nicholas J. Higham
There are modified versions of the Givens QR algorithm that use ``fast Givens rotations" which decrease the number of multiplications and square roots needed for the application of each Givens rotation.
% Fast Plane Rotations With Dynamic Scaling, Anda and Park, SIAM, 1994
It allows for the use of Givens rotations while requiring roughly the same number of floating point operations used by the Householder QR algorithm.
In practice, these methods are still not quite as fast as the Householder algorithm.
In spite of the need for more floating point operations, Givens rotations are still well suited for parallelization and can be much faster than the Householder algorithm when multiple processors are used.
% Givens and Householder Reductions for Linear Least Squares on a Cluster of Workstations, Omer Egecioglu and Ashok Srinivasan

\begin{problem}
\label{prob:Givens}
Write a function called \li{givens} that uses Givens rotations to compute the $QR$ decomposition of
a matrix $A$.
By performing successive Givens rotations, triangularize $A$ to find $R$.
Apply the rotations to an identity matrix as you go, then take the transpose to invert it and find $Q$.
Return $Q$ and $R$.
\end{problem}

We will now illustrate one example of how to use Givens rotations to change only specific parts of the array.

\begin{problem}
\label{prob:givens_hessenberg}
Write a modified version of your solution to Problem \ref{prob:Givens}, call it \li{givens2} that computes the 
QR decomposition of an upper Hessenberg matrix.
Instead of starting at the bottom of each column and working up, just run down the first subdiagonal from left to right.
When operating on $Q$, you do not need to operate on the full length of each row.
It is sufficient to perform the matrix multiplication on the portion of the array \li{Q[j:j+2,:j+2]} where \li{j} is the current column.

Note: You can zero out the portion below the first subdiagonal of a matrix like this:
\begin{lstlisting}
import numpy as np
import scipy.linalg as la
from numpy.random import rand
A = rand(500, 500)
A[1:] = la.triu(A[1:])
\end{lstlisting}

Notice how $Q$ in the QR decomposition of an upper Hessenberg matrix is also upper Hessenberg.
What is the computational order of  complexity for this problem?
Approximately for what $m$ is your implementation as fast as the general QR decomposition built in to \li{scipy.linalg} for computing the QR decomposition of an upper Hessenberg matrix?
\end{problem}

\begin{problem}
\label{prob:givens_hessenberg_modified}
You may have noticed that matrix multiplication by $Q$ is generally a $\mathcal{O} \left( n^3 \right)$ algorithm, while the application of these individual Givens rotations is a $\mathcal{O} \left( n^2 \right)$ algorithm.
Write a modified version of your solution to Problem \ref{prob:givens_hessenberg} called \li{givens2_mod} which returns 
an $(n-1) \times 2 \times 2$ array containing the computed values for $G$ at each step in the algorithm in the order in which they are applied to the upper Hessenberg array $H$.

Write two more functions \li{apply_Q} and \li{apply_QT} which, using the matrix of Givens rotations, perform left multiplication 
by $Q$ and $Q^{-1}$, respectively, on some other input array $B$.
Left multiplication by $Q^{-1}$ can be done by applying each of the Givens rotations to $B$ the same way you did to $H$ to compute its QR factorization.
Left multiplication by $Q$ can be done by applying the transpose of the Givens rotations to their corresponding portions of $B$, but in the reverse order.
Notice that you will have to apply each Givens rotation across the full width of the rows it operates on since you do not know anything about the content of $B$.

For around what size of matrices is direct multiplication by $Q$ slower than this method of multiplying by $Q$?
For timing purposes, make a random upper Hessenberg matrix, compute its QR decomposition using the function you just wrote and your solution to Problem \ref{prob:givens_hessenberg}, then time how long it takes to left multiply a random square array by $Q$ using the function you just wrote and the \li{dot} method of NumPy arrays.

Note: the functions you just wrote can be used to perform right multiplication as well since $B Q = \left(Q^T B^T \right)^T$ and $B Q^T = \left( Q B^T \right)^T$.
\end{problem}

An interesting side-note is that each Givens rotation can be represented as a single floating point number, so, when operating in place, $Q$ can be stored entirely in the lower triangular portion of the array on which we are operating by storing each rotation in the entry that it zeroes out.
A similar approach would to store Householder reflectors in the columns they zero out.
In either case, we can represent the QR decomposition of an array using only the memory that was originally used to store the array itself.
This is similar to the approach  for computing the LU decomposition entirely in place.
These representations of $Q$ and $R$ can be used in various ways to perform matrix multiplication by $Q$, $Q^T$ and $R$ as needed.

\section*{Least Squares}

We are now changing the subject to talk about line fitting.

It is well known that the displacement of a spring is proportional
to the force acting upon it, that is, $F = k x$.  The proportionality
constant $k$ is called Hooke's spring constant.  Consider a laboratory
experiment where different loads are placed on a spring and the displacement
is measured and recorded in the table below:
\vspace{5mm}\\
\begin{center}
\begin{tabular}{|c|c|}
	\hline
x & F \\
(cm) & (dyne)\\
\hline
1.04  & 3.11 \\
2.03  &  6.01\\
2.95  &  9.07\\
3.92  &  11.99\\
5.06  &  15.02\\
6.00  &  17.91\\
7.07  &  21.12\\
\hline
\end{tabular}
\end{center}
\vspace{5mm}
To find the spring constant $k$, we simply need to solve the following linear system
\[
\begin{pmatrix}
1.04\\
2.03\\
2.95\\
3.92\\
5.06\\
6.00\\
7.07\\
\end{pmatrix}
\begin{pmatrix}k\end{pmatrix} =
\begin{pmatrix}
3.11 \\
6.01\\
9.07\\
11.99\\
15.02\\
17.91\\
21.12\\
\end{pmatrix}.
\]
However, there is no solution to this system because it is overdetermined.
Instead, we seek the ``best'' $k$ that fits the data.
Least squares allows us to find that ``best'' solution.
We can find the least squares solution by computing the following in Python:
\begin{lstlisting}
>>> import numpy as np
>>> from scipy import linalg as la
>>> A = np.vstack([1.04,2.03,2.95,3.92,5.06,6.00,7.07])
>>> b = np.vstack([3.11,6.01,9.07,11.99,15.02,17.91,21.12])
>>> k = np.dot(np.dot(la.inv(np.dot(A.T,A)),A.T),b)
>>> k
array([[ 2.99568294]])
\end{lstlisting}
Hence, we find the spring constant to be $k = 2.9957$.
Note that \li{scipy.linalg} provides a built-in function for solving
least-squares problems.
We plot the data against the best fit as follows:
\begin{figure}[h!]
\label{fig1}
\begin{center}
\includegraphics[width=\textwidth]{line_lstsq}
\caption{The graph of the spring data together with its linear fit}
\label{Fig:SpringFit}
\end{center}
\end{figure}

\begin{lstlisting}
>>> from matplotlib import pyplot as plt
>>> x0 = np.linspace(0,8,100)
>>> y0 = k[0]*x0
>>> plt.plot(A,b,'*',x0,y0)
>>> plt.show()
\end{lstlisting}
See Figure \ref{Fig:SpringFit} to see how well the line fits the data.

\section*{General Line Fitting}

Suppose that we wish to fit a general line, that is $y=m x+b$, to the data set
$\{(x_k,y_k)\}^n_{k=1}$.  Assume that the line does not cross through the origin,
as in the previous example.  Then we seek both a slope and a $y$-intercept.
In this case, we set up the following linear system $A x = b$, or more precisely
\[
\begin{pmatrix}
x_1 & 1\\
x_2 & 1\\
x_3 & 1\\
\vdots & \vdots\\
x_n & 1
\end{pmatrix}
\begin{pmatrix}
m\\
b
\end{pmatrix}=
\begin{pmatrix}
y_1\\
y_2\\
y_3\\
\vdots\\
y_n
\end{pmatrix}.
\]
Note that $A$ has rank $2$ as long as not all of the $x_k$ values are the same.
Hence, the least squares solution
is given by
$$
\widehat{x} = (A^HA)^{-1}A^Hb.
$$
In what sense does this solution give us the best fit line for the data? Recall that since $A$ is injective,
the matrix $A(A^HA)^{-1}A^H$ is an orthogonal projector onto the range of $A$, which means that
$A(A^HA)^{-1}A^Hb = A\widehat{x}$ is the closest vector (with respect to the 2-norm) to $b$ that lies in the
range of $A$. That is, $\widehat{x}$ minimizes the error between $Ax$ and $b$, where the error is given
by the distance between these vectors, $\|b-Ax\|_2$. Another way to say this is that $\widehat{x}$ gives the
values $m$ and $b$ for which the sum of the squares of the distances from each data point $y_k$ to the value
$y = mx_k + b$ is as small as possible.

\section*{Loading Data from .npz Files}
For Least Squares problems as well as in many other contexts, loading data is often a necessary step before
proceeding with further analysis. Here we briefly review another data format in Python and the commands used
to load the data.

A \li{.npz} file is a compressed binary file that contains an archive of NumPy data structures.
A given file may therefore contain several arrays, each array associated with a unique string that identifies it.
When you load a \li{.npz} file in Python, a dictionary-like object is returned, and you can access the data by
providing the appropriate key. Note that when you load a \li{.npz} file, you must also be sure to close it when
you are finished. This is taken care of automatically if you use the \li{with ... as} keywords.

As an example, suppose that we have a file named \li{grades.npz} that contains several arrays, each giving the
homework scores of a particular student in a particular class. Assuming that one of the arrays is associated with
the key \li{'Abe'}, we can load this array in the following way:

\begin{lstlisting}
>>> with np.load('grades.npz') as grades:
>>>     abe_grades = grades['Abe']
>>> abe_grades
array([ 10.,  10.,  10.,  10.,  10.,  10.,  10.,  10.,  10.,  10.])
\end{lstlisting}

You will need to apply this technique in the next problem.

\begin{problem}
Write a function \li{fitLine} that takes no arguments and executes the following.
Load the \texttt{linepts} array from \texttt{data.npz}.
This consists of two columns corresponding to the $x$ and $y$ values of a given data set.
Use least squares to find the slope and $y$-intercept that best fits the data.
Then plot the data points and the line on the same graph.
The function should not return anything.
\end{problem}

\section*{Fitting data to a circle}

Recall that the equation of a circle, with radius $r$ centered at $(c_1,c_2)$, is given by
\begin{equation}
\label{circle}
(x-c_1)^2 + (y-c_2)^2 = r^2.
\end{equation}
Suppose we are given a set of data points closely forming a circle $\{(x_i,y_i)\}^n_{i=1}$.
The ``best'' fit is found via least squares by expanding \eqref{circle} to get
\[
2 c_1 x + 2 c_2 y + c_3 = x^2 + y^2,
\]
where $c_3 = r^2 - c_1^2 - c_2^2$.  Then we can write the linear system $A x = b$ as
\[
\begin{pmatrix}
2 x_1 & 2 y_1 & 1\\
2 x_2 & 2 y_2 & 1\\
\vdots & \vdots & \vdots \\
2 x_n & 2 y_n & 1
\end{pmatrix}
\begin{pmatrix}
c_1\\
c_2\\
c_3
\end{pmatrix}=
\begin{pmatrix}
x_1^2 + y_1^2\\
x_2^2 + y_2^2\\
\vdots\\
x_n^2 + y_n^2
\end{pmatrix},
\]
where the matrix $A$ and the vector $b$ are obtained by the given data and the unknown
$x$ contains the information about the center and radius of the circle and is obtained
by finding the least squares solution.

\section*{Example}

In this section, we fit the following points to a circle:
\begin{align*}
&(134,76),(104,146),(34,176),(-36,146),\\
&(-66,76),(-36,5),(34,-24),(104,5),(134,76)
\end{align*}

We enter them into Python as a $9\times 2$ array:
\begin{lstlisting}
>>> P = np.array([[134,76],[104,146],[34,176],[-36,146],
                  [-66,76],[-36,5],[34,-24],[104,5],[134,76]])
\end{lstlisting}
We compute $A$ and $b$ by entering the following:
\begin{lstlisting}
>>> A = A =np.hstack((2*P, np.ones((9,1))))
>>> b = (P**2).sum(axis=1)
\end{lstlisting}
Hence, we get the least squares solution
\begin{lstlisting}
>>> x = np.dot(np.dot(la.inv(np.dot(A.T,A)),A.T),b)
\end{lstlisting}
Then we find $c_1$, $c_2$, and $r$ by:
\begin{lstlisting}
>>> from math import sqrt
>>> c1, c2, c3 = x
>>> r = sqrt(c1**2 + c2**2 + c3)
\end{lstlisting}
We plot this by executing
\begin{lstlisting}
>>> theta = np.linspace(0,2*np.pi,200)
>>> plt.plot(r*np.cos(theta)+c1,r*np.sin(theta)+c2,'-',P[:,0],P[:,1],'*')
>>> plt.show()
\end{lstlisting}

\begin{problem}
Write a function \li{fitCircle} that does the following.
Load the \texttt{circlepts} array from \texttt{data.npz}.
This consists of two columns corresponding to the $x$ and $y$ values of a given
data set.  Use least squares to find the center and radius of the circle that best
fits the data.  Then plot the data points and the circle on the same graph.
The function should return nothing.
\end{problem}

\begin{problem}
The general equation for an ellipse is:
\[
ax^2 + bx + cxy + dy + ey^2 = 1
\]

Write a function \li{fitEllipse} that uses least squares to fit data to an ellipse.
The function should take a $n\times 2$ array as input, where the first column gives the $x$-coordinates
and the second column gives the $y$-coordinates. Find the least squares solution for $a, b, c, d,$ and $e$,
and return the solution.
You can test out your function on the \texttt{ellipsepts} array from \texttt{data.npz}. You should get  $0.087$, $-0.141$,  $0.159$, $-0.316$, $0.366$ for $a, b, c, d,$ and $e$ respectively.
\end{problem}

In these Least Squares problems, we have found best fit lines and ellipses relative to the 2-norm.
It is possible to generalize the idea of best fit curves relative to other norms.
See Figure \ref{Fig:ellipse} for an illustration of this.

\begin{figure}[h]
\label{ellipsefit}
\centering
\includegraphics[width=\textwidth]{ellipsefit.pdf}
\caption{Fitting an ellipse using different norms.}
\label{Fig:ellipse}
\end{figure}

### 2014-07-19 Algorithms/MST/mst.tex
Minor edits to clarify the lab better

\lab{Algorithm}{Kruskal's and Prim's Algorithm}{Kruskal's and Prim's Algorithm}
\label{Ch:Kruskal}

\objective{Find a minimum spanning tree for a connected, weighted graph using Kruskal's Algorithm and Prim's Algorithm. Apply these techniques to segment an image.}

\section*{Weighted Graphs and Spanning trees}

Remember that a graph is composed of two sets: a set of nodes and a set of edges that connect the nodes.

\begin{figure}[H]
\includegraphics[width = .4\textwidth]{graph1.pdf}
\caption{An example of an undirected graph}
\label{mst:graph1}
\end{figure}

A graph is \emph{undirected} if its edges are bi-directional. This means that the edges have no specific direction in between nodes. Figure
\ref{mst:graph1} shows an example of an undirected graph.
A \emph{weighted graph} is a graph where each edge has a value associated with it.
Usually these values represent some sort of cost or distance.
A \emph{connected graph} is a graph where there is a path, or a set of edges, that connects every two nodes together.

We can represent a connected graph by a matrix.
Each row of the matrix represents a starting point and each column represents a destination.
If an edge from node \emph{i} to node \emph{j} exists, we put the weight of the edge in the \emph{i},\emph{j} entry of the matrix.
If there is no edge, we put a 0.
For the above graph in Figure \ref{mst:graph1} we generate the following matrix:

\[
A = \begin{pmatrix}
0 & 1 & 0 & 0 & 0 & 0\\
1 & 0 & 1 & 0 & 0 & 1\\
0 & 1 & 0 & 1 & 1 & 1\\
0 & 0 & 1 & 0 & 1 & 0\\
0 & 0 & 1 & 1 & 0 & 1\\
0 & 1 & 1 & 0 & 1 & 0\\
\end{pmatrix}
\]

This matrix is called an \emph{adjacency matrix}.
Note that the graph is undirected in this example and therefore the adjacency matrix is symmetric.

Now consider the graph in Figure \ref{mst:graph4}.  This graph is similar to the graph in Figure \ref{mst:graph1}, except now weights are attached to each edge.  
\begin{figure}[H]
\includegraphics[width = .4\textwidth]{graph4.pdf}
\caption{A weighted, undirected graph}
\label{mst:graph4}
\end{figure}

The adjacency matrix for this graph is 
\[
A = \begin{pmatrix}
0 & 3 & 0 & 0 & 0 & 6\\
3 & 0 & 5 & 0 & 0 & 4\\
0 & 5 & 0 & 1 & 1 & 5\\
0 & 0 & 1 & 0 & 2 & 0\\
0 & 0 & 1 & 2 & 0 & 4\\
6 & 4 & 5 & 0 & 4 & 0\\
\end{pmatrix}
\]

Another way to store the information from a graph is to make a list of the edges with their corresponding weights.
To do this for an unweighted, undirected graph, we simply create a list of the pairs of nodes that correspond to each edge.
For a weighted, undirected graph, a third value could be added to each pair of nodes to represent the corresponding weight of the edge.
A list like this for the graph in Figure \ref{mst:graph1} would look like this:

\begin{align*}
[('A', 'B'),
 ('B', 'C'),
 ('B', 'F'),
 ('C', 'D'),\\
 ('C', 'E'),
 ('C', 'F'),
 ('D', 'E'),
 ('E', 'F')]
\end{align*}

A \emph{spanning tree} of a connected, undirected graph $G$ is an undirected graph that contains all the nodes of $G$, a subset of the edges, and no cycles.
A cycle, for undirected graphs, is a path that starts and ends on the same node without crossing any edge more than once. Figure \ref{mst:graph3} shows an example of an undirected graph with a cycle in it. The cycle includes the nodes A, B, C, and F.

\begin{figure}[H]
\includegraphics[width = .4\textwidth]{graph3.pdf}
\caption{A cycle in an undirected graph}
\label{mst:graph3}
\end{figure}

A minimum spanning tree (MST) of a weighted, undirected graph is a spanning tree where the total weight is less than or equal to the total weight of every other spanning tree.
Both Kruskal's and Prim's Algorithms are methods that find a minimum spanning tree of a weighted, undirected graph.
Figure \ref{mst:graph2} shows a spanning tree of the graph shown in Figure \ref{mst:graph1}.
Figure 5.5 shows a minimum spanning tree of the graph shown in Figure \ref{mst:graph4}.

\begin{figure}[H]
\includegraphics[width = .4\textwidth]{graph2.pdf}
\caption{A spanning tree for the graph in Figure \ref{mst:graph1}.}
\label{mst:graph2}
\end{figure}

\begin{figure}[H]
\includegraphics[width = .4\textwidth]{graph5.pdf}
\caption{A MST of the graph in Figure \ref{mst:graph4}.}
\label{mst:graph5}
\end{figure}

\section*{Kruskal's algorithm}

Given a weighted, directed graph G with $n$ nodes, Kruskal's algorithm finds a minimum spanning tree by first sorting the edges from smallest to largest.
Then starting with the smallest, the algorithm adds edges to the tree as long as the addition of each new edge does not create a cycle.
When $n-1$ edges have been added, the algorithm stops.

In order to avoid creating cycles while building the tree, it is necessary to keep track of which portions of the tree currently lie in connected groups.
This can be done by creating a dictionary where, when the algorithm starts, each node points to itself as the ``root" of its own tree.
As we add edges to the tree we will change which nodes are the ``root" nodes. This tracks which nodes are currently connected.
Doing this allows us to run the algorithm by iterating over the edges by weight in ascending order, and adding them to the tree if they connect nodes which are not already connected by the current tree.

We will consider the graph in Figure \ref{mst:graph4} and apply Kruskal's algorithm.
First, we initialize our spanning tree to be an empty list. Then, to track the root nodes of each tree, we initialize a dictionary where each node points to itself.

\begin{lstlisting}
spanning_tree = []
nodes = {'A': 'A', 'B':'B', 'C':'C', 'D':'D', 'E':'E', 'F':'F'}
\end{lstlisting}
Since there are 6 nodes in the graph, we will continue until we have 5 edges in the tree.
Next, sort the edges by weight.

\begin{lstlisting}
edges = [(C, D, 1), (C, E, 1), (D, E, 2), (A, B, 3), (B, F, 4), (E, F, 4), 
         (B, C, 5), (C, F, 5), (A, F, 6)]
\end{lstlisting}

Now we begin iterating through the edges to build the tree.
The first edge in the list is \li{(C,D,1)}.
The root for \li{C} is \li{C} and the root for \li{D} is \li{D}, so we add this edge to the tree.
The tree becomes \li{[(C, D, 1)]}.
We then change the root node of \li{D} to be \li{C}.
The dictionary of root nodes now looks like

\begin{lstlisting}
{'A': 'A', 'B':'B', 'C':'C', 'D':'C', 'E':'E', 'F':'F'}
\end{lstlisting}

Now we process the next edge, \li{(C, E, 1)}.
The root node of \li{C} is \li{C}, and the root node of \li{E} is \li{E}, so adding this edge does not create a cycle.
We add this edge into the tree, so the tree is now \li{[(C, D, 1), (C, E, 1)]}.
Then we change the root of \li{E} so that it is \li{C}, so the dictionary is

\begin{lstlisting}
{'A': 'A', 'B':'B', 'C':'C', 'D':'C', 'E':'C', 'F':'F'}
\end{lstlisting}

The next edge is \li{(D, E, 2)}.
The root node of the tree containing \li{D} is \li{C} and the root node of the tree containing \li{E} is \li{C}, so these nodes are already connected, so we do not add this edge to the tree.

The next edge is \li{(A, B, 3)}.
The root node for \li{A} is \li{A}, and the root node for \li{B} is \li{B}, so we add the edge to the tree and update the dictionary.
The dictionary becomes

\begin{lstlisting}
{'A': 'A', 'B':'A', 'C':'C', 'D':'C', 'E':'C', 'F':'F'}
\end{lstlisting}

The next edge is \li{(B, F, 4)}.
The root node for \li{B} is \li{A} and the root node for \li{F} is \li{F}, so we add the edge to the tree and change the root node of \li{F} to be \li{A}.
The dictionary becomes

\begin{lstlisting}
{'A': 'A', 'B':'A', 'C':'C', 'D':'C', 'E':'C', 'F':'A'}
\end{lstlisting}

The next edge is \li{(E, F, 4)}.
The root node for \li{E} is \li{C} and the root node for \li{F} is \li{A}.
We add the edge to the tree and end the algorithm since there are now 5 edges in the tree.
(If we were to continue the algorithm, we would update the root node of \li{A} to be \li{C}.)

Notice how we updated the root of the root node to one of the nodes in the edge added.
We manage the root nodes in this manner in order to find the root node by tracing back through the dictionary until we find a node that points to itself.
For example, to find the root node of \li{F} we first need to get the value for \li{F} from the dictionary.  This is \li{A}.
Then we get the value for \li{A} from the dictionary which is \li{C}.
Since the value of \li{C} in the dictionary is itself, the \li{C} is the root node of the graph containing \li{F}.
It is necessary to trace through the dictionary in this manner each time because we avoid iterating over all of the nodes and updating their root values everytime.

Below is the pseudocode for the algorithm as we have described it above:

\begin{itemize}

% Note: The comments in the solutions match the pseudocode.
% When making changes, please keep that in mind.

\item Initialize an empty list of edges for the minimum spanning tree.

\item Make a dictionary that points each node toward its root (not always directly to it).
Start with each node pointing to itself.

\item Initialize the number of nodes that still need to be processed to the number of nodes minus 1.

\item Define a helper function that, given a node, traces through the dictionary to find the root of its tree.
This can be done like this:

	\begin{itemize}

	\item Initialize a temporary variable to be the node for which we are finding the root.

	\item While the temporary node does not point to itself in the dictionary:

		\begin{itemize}

		\item Update the temporary node to be the node it currently points to in the dictionary.

		\end{itemize}

	\item Return the temporary node.

	\end{itemize}

\item Iterate over the edges by ascending weight.
Use a \li{for} loop for this and return the tree when it is big enough which breaks the loop for you.

	\begin{itemize}

	\item Trace through the dictionary to find the root node of each of the nodes in the edge you are processing.

	\item If the roots are not the same (i.e. if adding the edge doesn't form a cycle):

		\begin{itemize}

		\item Add the edge to the tree.

		\item Lower the number of edges remaining by one.

		\item If the number of edges remaining is 0, return the tree (which also breaks the loop).

		\item Update the root of the root of the second node in the edge to be the root of the first node in the edge.
			This lets us record that the two subtrees are now connected.

		\end{itemize}

	\end{itemize}

\end{itemize}
Note on implementation: You can iterate over a sorted copy of a list using the built in \li{sorted} function.
You can sort by the third value in each tuple using the \li{itemgetter} function that is part of the \li{operator} library included with Python.
For example:
\begin{lstlisting}
from operator import itemgetter
...
for n1, n2, weight in sorted(edges, key=itemgetter(2)):
    ...
\end{lstlisting}

\begin{problem}
Implement Kruskal's algorithm.
Test your algorithm on random symmetric arrays.
You can generate a random symmetric array by multiplying a random array by its transpose.
Use the data from MSTdata.npy to test your tree.
Use np.load("MSTdata.npy") to get it.
Use the \li{formChanger} function below to put it in the right form.
\begin{lstlisting}
def formChanger(oldData):
    newData=[]
    for i in oldData: newData.append((i[0],i[1],int(i[2])))
    return newData
\end{lstlisting}
\end{problem}
\section*{Prim's algorithm}

Prim's is a similar algorithm for finding minimum spanning trees.
While it is much slower than Kruskal's algorithm for sparse graphs, it is much faster for dense graphs because Prim's algorithm avoids sorting the edges.

Again, consider the example shown in Figure 7.4.
We first initialize a dictionary with all the nodes as keys in order to track which nodes have not been processed.
At the beginning it will be \li{\{A:False, B:False, C:False, D:False, E:False, F:False\}}.
We then form the dictionary that maps each node to the edges that contain it.
Since we will already know one node of each edge while we use the dictionary, we only need to store the node that is not being looked up.
It should end up looking like \li{\{A:[(B, 3), (F, 6)], B:[(A, 3), (C, 5), (F, 4)], C:[(B, 5), (D, 1), (E, 1), (F, 5)], D:[(C, 1), (E, 2)], E:[(C, 1), (D, 2), (F, 4)], F:[(A, 6), (B, 4), (C, 5), (E, 4)]\}}.
We will also initialize an empty dictionary to track the shortest edges that run between nodes we have processed and nodes that we haven't.
As we iterated over the edges in our initialization step, we are also able to find the shortest edge.
In this case, it is \li{(D, C, 1)}.
Let's start with \li{(D, C, 1)} and initialize our tree as the list \li{[(D, C, 1)]}.
Next, we mark \li{D} and \li{C} as processed in the dictionary that tracks which nodes have been processed.
We now start to build our dictionary of nodes that are one edge away from our processed nodes.
In this case, after adding the shortest edges between processed and unprocessed nodes to the dictionary, the dictionary becomes \li{\{B:(C, B, 5), E:(C, E, 1), F:(C, F, 5)\}}.
Notice we did not include \li{E:(D, E, 2)} because there is a shorter edge to \li{E} from \li{C}.

Of the edges in the dictionary of edges that can be processed next, the shortest is \li{(C, E, 1)}, so we add that edge to the tree and mark \li{E} as processed.
With \li{E} being processed, we can now reach \li{F} at a cost of 4.
After making this change, the dictionary of edges to process becomes \li{\{B:(C, B, 5), E:(C, E, 1), F:(E, F, 4)\}}.
Since \li{E} no longer needs to be processed, we can remove it from consideration.
So this dictionary now becomes \li{\{B:(C, B, 5), F:(E, F, 4)\}}.

Of the edges to be processed next, \li{(E, F, 4)} is the shortest, so we add it to the tree and mark \li{F} as processed.
After performing the appropriate modifications to the dictionary of edges to be processed next, it becomes \li{\{B:(F, B, 4), A:(F, A, 6)\}}.

Of the edges to be processed next, \li{(F, B, 4)} is the shortest, so we add it to the tree and mark \li{B} as processed.
After performing the appropriate modifications to the dictionary of edges to be processed next, it becomes \li{\{A:(B, A, 3)\}}.

The only edge to be considered is \li{(B, A, 3)}, so we add it to the tree.
The tree is long enough that it it spans the nodes, so the algorithm is finished.

Here's pseudocode for a version of Prim's algorithm.
While it is not a perfectly optimized version, it is pretty good.

\begin{itemize}

% Note: The comments in the solutions match the pseudocode.
% When making changes, pleas keep that in mind.

\item Initialize a dictionary to track which nodes have been processed.

\item Initialize an empty dictionary of lists to track the edges containing each node.

\item Fill the edge list.
	Be sure to add each edge to the list corresponding to both of its nodes.

\item Get the first edge to add (The shortest edge from any given node is a good pick).

\item Mark the nodes in the first edge as processed.

\item Initialize the tree to be the list containing the first edge.

\item Initialize an empty dictionary that will be used to contain the edges that can be processed next.

\item Define a helper function to insert an edge into the dictionary (if that insertion is needed).
	This can be done as follows:

	\begin{itemize}

	\item Get the value of the node that is reached by the edge.

	\item If that node isn't in the dictionary, set its value to be the edge passed to the functions.

	\item If it is in the dictionary already, set its value to be the shorter of the edges being processed and the edges already in the dictionary.

	\end{itemize}

\item Use the helper function to insert the edges reached by the first two processed nodes into the dictionary of edges to be processed.

\item Until the tree contains enough edges to span all the nodes:

	\begin{itemize}

	\item Find the shortest edge in the dictionary of edges to be processed.

	\item Remove the shortest edge from the dictionary.

	\item Add it to the tree.

	\item Mark the node reached by the new edge as processed.

	\item Use the helper function to insert the edges reached by the newly processed node into the dictionary of edges to be processed.

	\end{itemize}

\item Return the completed tree.

\end{itemize}

\begin{problem}
Write a Python function that uses Prim's algorithm to find the minimum spanning tree of a graph.
Test your implementation with the same data as the previous problem.
Compare the speed of Prim's algorithm with the speed of Kruskal's algorithm.
Create a function which prints these two times.
\end{problem}

Many graph packages have functions for finding the minimal spanning tree. One of these is NetworkX. You begin as follows
\begin{lstlisting}
import networkx as nx
G=nx.Graph()
\end{lstlisting}
You can add a node $x$ with \li{G.add_node(x)} and add a edge from $x$ to $y$ with weight n with \li{G.add_edge(x,y,weight=n)}. Then \li{nx.minimum_spanning_tree(G)} returns the minimal spanning tree using Kruskals algorithm.

\begin{problem}
Use NetworkX to find the minimal spanning tree of the data in MSTdata.npy.
Compare the output with that of your algorithm.
Create a function which prints your comparison.
\end{problem}

%Add specifications

\section*{Image Segmentation}

%Lab \ref{MSTImgSeg}

One application of Minimal Spanning Trees (MSTs) is image segmentation.
Kruskal's algorithm is especially good at this.
You can convert an image into graph. Each pixel is a vertex and you define weights between the nodes. You then build the minimal spanning tree. If you take away the edge with the greatest weight you will get two forests (removing an edge from any graph with no cycles will create two forests). Then you can continue doing this to each forest to split up those forests. Each forest corresponds to a segment of the image.
Let $k$ be the number of divisions that is wanted and $n$ be the number of nodes.
Kruskal's algorithm is performed until $n-(k+1)$ edges are added which is the same as taking out the $k$ edges with greatest weights.

There are many different ways to turn an image into a graph and weight the edges.
A simple, yet effective, version is to make every pixel a node and the edges are the difference in intensities in the four cardinal directions.

This means that there are less than $4n$ edges.
Other image segmentation algorithms have to use $n^2$ space.
This gives the MST algorithm a critical advantage over other image segmentation algorithms.

\begin{problem}
Write a function that takes a black and white image as input and outputs a list of the edges and a list of nodes.
Store the edges using the form \li{(node,node,weight)}.
\end{problem}

The provided Kruskal's algorithm takes as inputs the list of nodes, the list of edges and the number of divisions desired.
The number of divisions often has to be higher than the actual number that is needed because sometimes one or two pixels form a division because the difference between them and the pixels around them is so great.
You will have to adjust the number of divisions until the desired result is found.  See Figure 9.1.

\begin{problem}
Perform the image segmentation algorithm on the image, then graph the original image and the three largest divisions.
(Use the Counter class from collections to find the number of pixels in each division.)
\end{problem}

\begin{problem}
Make a division of the image a different color.
\end{problem}

\vfill
\begin{figure}[ht]
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg1.jpg}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg2.jpg}
\end{minipage}
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg3.jpg}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg4.jpg}
\end{minipage}
\caption{The original image is in the top left hand corner. The three larges segments are shown in the other corners. The original image was 498x498 and 50000 divisions were used.}
\end{figure}
\vfill

### 2014-07-20 Algorithms/MST/mst.tex
Reworded some paragraphs and added some code for clarification

@@
 
-Given a weighted, directed graph G with $n$ nodes, Kruskal's algorithm finds a minimum spanning tree by first sorting the edges from smallest to largest.
+Given a weighted, directed graph G with $n$ nodes, Kruskal's algorithm finds a minimum spanning tree by first sorting the edges by weight from smallest to largest.
 Then starting with the smallest, the algorithm adds edges to the tree as long as the addition of each new edge does not create a cycle.
@@
 
-In order to avoid creating cycles while building the tree, it is necessary to keep track of which portions of the tree currently lie in connected groups.
+In order to avoid creating cycles while building the spanning tree, it is necessary to keep track of which portions of the tree currently lie in connected groups.
 This can be done by creating a dictionary where, when the algorithm starts, each node points to itself as the ``root" of its own tree.
-As we add edges to the tree we will change which nodes are the ``root" nodes. This tracks which nodes are currently connected.
-Doing this allows us to run the algorithm by iterating over the edges by weight in ascending order, and adding them to the tree if they connect nodes which are not already connected by the current tree.
+As we add edges to the spanning tree we will change which nodes are the ``root" nodes. This tracks which nodes are currently connected.
+Doing this allows us to add edges to the spanning tree if they connect nodes which are not already connected by the current spanning tree.
 
 We will consider the graph in Figure \ref{mst:graph4} and apply Kruskal's algorithm.
-First, we initialize our spanning tree to be an empty list. Then, to track the root nodes of each tree, we initialize a dictionary where each node points to itself.
+First, we initialize our spanning tree to be an empty list. Then, to track the root nodes of each connected tree, we initialize a dictionary where each node points to itself.
 
@@
 \end{lstlisting}
-Since there are 6 nodes in the graph, we will continue until we have 5 edges in the tree.
+Since there are six nodes in the graph, we will continue until we have five edges in the spanning tree.
 Next, sort the edges by weight.
@@
 
-Now we begin iterating through the edges to build the tree.
+Now we begin iterating through the edges to build the spanning tree.
 The first edge in the list is \li{(C,D,1)}.
 The root for \li{C} is \li{C} and the root for \li{D} is \li{D}, so we add this edge to the tree.
-The tree becomes \li{[(C, D, 1)]}.
+The tree becomes
+\begin{lstlisting}
+[('C', 'D', 1)]
+\end{lstlisting}
 We then change the root node of \li{D} to be \li{C}.
@@
 The root node of \li{C} is \li{C}, and the root node of \li{E} is \li{E}, so adding this edge does not create a cycle.
-We add this edge into the tree, so the tree is now \li{[(C, D, 1), (C, E, 1)]}.
-Then we change the root of \li{E} so that it is \li{C}, so the dictionary is
+We add this edge into the tree, so the tree is now
+
+\begin{lstlisting}
+[('C', 'D', 1), ('C', 'E', 1)]
+\end{lstlisting}
+Then we change the root of \li{E} so that it is \li{C}. The dictionary is now
 
@@
 The next edge is \li{(D, E, 2)}.
-The root node of the tree containing \li{D} is \li{C} and the root node of the tree containing \li{E} is \li{C}, so these nodes are already connected, so we do not add this edge to the tree.
+The root node of the tree containing \li{D} is \li{C} and the root node of the tree containing \li{E} is \li{C}, so these nodes are already connected, so we do not add this edge to the spanning tree.
 
@@
 \end{lstlisting}
+The tree becomes
+\begin{lstlisting}
+[('C', 'D', 1), ('C', 'E', 1), ('A', 'B', 3)]
+\end{lstlisting}
 
@@
 The root node for \li{B} is \li{A} and the root node for \li{F} is \li{F}, so we add the edge to the tree and change the root node of \li{F} to be \li{A}.
+
 The dictionary becomes
@@
 \end{lstlisting}
-
+The tree becomes
+\begin{lstlisting}
+[('C', 'D', 1), ('C', 'E', 1), ('A', 'B', 3), ('B', 'F', 4)]
+\end{lstlisting}
 The next edge is \li{(E, F, 4)}.
 The root node for \li{E} is \li{C} and the root node for \li{F} is \li{A}.
-We add the edge to the tree and end the algorithm since there are now 5 edges in the tree.
-(If we were to continue the algorithm, we would update the root node of \li{A} to be \li{C}.)
+We add the edge to the tree and end the algorithm since the spanning tree now has five edges.
 
-Notice how we updated the root of the root node to one of the nodes in the edge added.
-We manage the root nodes in this manner in order to find the root node by tracing back through the dictionary until we find a node that points to itself.
+The final spanning tree looks like
+\begin{lstlisting}
+[('C', 'D', 1), ('C', 'E', 1), ('A', 'B', 3), ('B', 'F', 4), ('E', 'F', 4)]
+\end{lstlisting}
+If we were to continue the algorithm, we would update the root node of \li{A} to be \li{C}. The dictionary would then look like
+
+\begin{lstlisting}
+{'A': 'C', 'C': 'C', 'B': 'A', 'E': 'C', 'D': 'C', 'F': 'A'}
+\end{lstlisting}
+
+One of the keys to this algorithm is to make sure we are not adding cycles into the spanning tree.
+A way to think about how the algorithm does this is by picturing it building smaller trees that have a common root node, and then as the algorithm progresses, all of these smaller trees join to have the same root node.
+To find the root node of the tree, we look through our dictionary until we find a node that points to itself.
 For example, to find the root node of \li{F} we first need to get the value for \li{F} from the dictionary.  This is \li{A}.
 Then we get the value for \li{A} from the dictionary which is \li{C}.
-Since the value of \li{C} in the dictionary is itself, the \li{C} is the root node of the graph containing \li{F}.
+Since the value of \li{C} in the dictionary is itself, then \li{C} is the root node of the graph containing \li{F}.
 It is necessary to trace through the dictionary in this manner each time because we avoid iterating over all of the nodes and updating their root values everytime.
 
-Below is the pseudocode for the algorithm as we have described it above:
+Below is the outline of the algorithm we have described above:
 

### 2014-07-21 Algorithms/Givens/Givens.tex
minor edits to givens lab

@@
 Because Givens rotations only operate on two specific rows at a time, they provide multiple ways to iterate over the array we are processing.
-For example, using Givens rotations you could, starting at the bottom of each column and continuing up until you reach the main diagonal, zero out each entry by applying a Givens rotation to each row and the row immediately above it.
-Alternatively, within each column, you could could zero out the first entry of each row below the main diagonal by rotating it with the row corresponding to the main diagonal.
-Yet, another way to do it would be to start at the bottom left corner of the matrix and then zero out each entry from left to right along each diagonal.
-This flexibility is what makes the use of Givens rotations ideal for some problems.
-When working with sparse matrices, Givens rotations allow for the solution of linear systems through a series of orthonormal operations that operate only on small parts of the array.
+Here are three examples showcasing different iterating processes:
+
+\begin{enumerate}
+\item Starting at the bottom of each column and continuing up until you reach the main diagonal, zero out each entry by applying a Givens rotation to each row and the row immediately above it.
+
+\item Starting within each column, zero out the first entry of each row below the main diagonal by rotating it with the row corresponding to the main diagonal.
+
+\item Starting at the bottom left corner of the matrix, zero out each entry from left to right along each diagonal.
+
+\end{enumerate}
+
+This flexibility is what makes using Givens rotations ideal for some problems.
+When working with sparse matrices, Givens rotations allow for the solution of linear systems through a series of orthonormal operations that operate only on small portions of the array.
 The fact that Givens rotations operate only on small portions of the array also lends well to parallel solutions to systems of equations (or least squares problems).
@@
 \label{prob:givens_hessenberg}
-Write a modified version of your solution to Problem \ref{prob:Givens}, call it \li{givens2} that computes the 
-QR decomposition of an upper Hessenberg matrix.
+Write a modified version of your solution to Problem \ref{prob:Givens} called \li{givens2} that computes the QR decomposition of an upper Hessenberg matrix.
 Instead of starting at the bottom of each column and working up, just run down the first subdiagonal from left to right.
@@
 What is the computational order of  complexity for this problem?
-Approximately for what $m$ is your implementation as fast as the general QR decomposition built in to \li{scipy.linalg} for computing the QR decomposition of an upper Hessenberg matrix?
+For approximately what $m$ is your implementation as fast as the general QR decomposition built into \li{scipy.linalg} for computing the QR decomposition of an upper Hessenberg matrix?
 \end{problem}
@@
 \label{prob:givens_hessenberg_modified}
-You may have noticed that matrix multiplication by $Q$ is generally a $\mathcal{O} \left( n^3 \right)$ algorithm, while the application of these individual Givens rotations is a $\mathcal{O} \left( n^2 \right)$ algorithm.
-Write a modified version of your solution to Problem \ref{prob:givens_hessenberg} called \li{givens2_mod} which returns 
+You may have noticed that matrix multiplication by $Q$ is generally a $\mathcal{O} \left( n^3 \right)$ algorithm even though the application of these individual Givens rotations is a $\mathcal{O} \left( n^2 \right)$ algorithm.
+
+Now write a modified version of your solution to Problem \ref{prob:givens_hessenberg} called \li{givens2_mod} which returns 
 an $(n-1) \times 2 \times 2$ array containing the computed values for $G$ at each step in the algorithm in the order in which they are applied to the upper Hessenberg array $H$.
+\end{problem}
 
+\begin{problem}
 Write two more functions \li{apply_Q} and \li{apply_QT} which, using the matrix of Givens rotations, perform left multiplication 
@@
 An interesting side-note is that each Givens rotation can be represented as a single floating point number, so, when operating in place, $Q$ can be stored entirely in the lower triangular portion of the array on which we are operating by storing each rotation in the entry that it zeroes out.
-A similar approach would to store Householder reflectors in the columns they zero out.
+A similar approach would be to store Householder reflectors in the columns they zero out.
 In either case, we can represent the QR decomposition of an array using only the memory that was originally used to store the array itself.
 This is similar to the approach  for computing the LU decomposition entirely in place.
-These representations of $Q$ and $R$ can be used in various ways to perform matrix multiplication by $Q$, $Q^T$ and $R$ as needed.
+These representations of $Q$ and $R$ can be used in various ways to perform matrix multiplication by $Q$, $Q^T$, and $R$ as needed.
 
@@
 
-We are now changing the subject to talk about line fitting.
+Let's now talk about line fitting.
 
@@
 \section*{Loading Data from .npz Files}
-For Least Squares problems as well as in many other contexts, loading data is often a necessary step before
-proceeding with further analysis. Here we briefly review another data format in Python and the commands used
-to load the data.
+For Least Squares problems as well as in many other contexts, loading data is often a necessary step before proceeding with further analysis.
+We will briefly review another data format in Python and the commands used to load the data.
 
@@
 A given file may therefore contain several arrays, each array associated with a unique string that identifies it.
-When you load a \li{.npz} file in Python, a dictionary-like object is returned, and you can access the data by
-providing the appropriate key. Note that when you load a \li{.npz} file, you must also be sure to close it when
+When you load a \li{.npz} file in Python, a dictionary-like object is returned and you can access the data by providing the appropriate key.
+Note that when you load a \li{.npz} file, you must also be sure to close it when
 you are finished. This is taken care of automatically if you use the \li{with ... as} keywords.
@@
 \begin{lstlisting}
->>> A = A =np.hstack((2*P, np.ones((9,1))))
+>>> A = np.hstack((2*P, np.ones((9,1))))
 >>> b = (P**2).sum(axis=1)

### 2014-07-26 Algorithms/MST/mst.tex
Finished up Kruskal's and Prim's section

@@
 \begin{lstlisting}
-edges = [(C, D, 1), (C, E, 1), (D, E, 2), (A, B, 3), (B, F, 4), (E, F, 4), 
-         (B, C, 5), (C, F, 5), (A, F, 6)]
+edges = [('C', 'D', 1), ('C', 'E', 1), ('D', 'E', 2), ('A', 'B', 3), 
+	     ('B', 'F', 4), ('E', 'F', 4), ('B', 'C', 5), ('C', 'F', 5), 
+	     ('A', 'F', 6)]
 \end{lstlisting}
@@
 Now we begin iterating through the edges to build the spanning tree.
-The first edge in the list is \li{(C,D,1)}.
-The root for \li{C} is \li{C} and the root for \li{D} is \li{D}, so we add this edge to the tree.
-The tree becomes
+The first edge in the list is \li{('C', 'D',1)}.
+The root for \li{'C'} is \li{'C'} and the root for \li{'D'} is \li{'D'}, so we add this edge to the tree.
+The tree becomes:
 \begin{lstlisting}
@@
 \end{lstlisting}
-We then change the root node of \li{D} to be \li{C}.
-The dictionary of root nodes now looks like
+We then change the root node of \li{'D'} to be \li{'C'}.
+The dictionary of root nodes now looks like:
 
@@
 
-Now we process the next edge, \li{(C, E, 1)}.
-The root node of \li{C} is \li{C}, and the root node of \li{E} is \li{E}, so adding this edge does not create a cycle.
-We add this edge into the tree, so the tree is now
+Now we process the next edge, \li{('C', 'E', 1)}.
+The root node of \li{'C'} is \li{'C'}, and the root node of \li{'E'} is \li{'E'}, so adding this edge does not create a cycle.
+We add this edge into the tree, so the tree is now:
 
@@
 \end{lstlisting}
-Then we change the root of \li{E} so that it is \li{C}. The dictionary is now
+Then we change the root of \li{'E'} so that it is \li{'C'}. The dictionary is now:
 
@@
 
-The next edge is \li{(D, E, 2)}.
-The root node of the tree containing \li{D} is \li{C} and the root node of the tree containing \li{E} is \li{C}, so these nodes are already connected, so we do not add this edge to the spanning tree.
+The next edge is \li{('D', 'E', 2)}.
+The root node of the tree containing \li{'D'} is \li{'C'} and the root node of the tree containing \li{'E'} is \li{'C'}, so these nodes are already connected, so we do not add this edge to the spanning tree.
 
-The next edge is \li{(A, B, 3)}.
-The root node for \li{A} is \li{A}, and the root node for \li{B} is \li{B}, so we add the edge to the tree and update the dictionary.
-The dictionary becomes
+The next edge is \li{('A', 'B', 3)}.
+The root node for \li{'A'} is \li{'A'}, and the root node for \li{'B'} is \li{'B'}, so we add the edge to the tree and update the dictionary.
+The dictionary becomes:
 
@@
 \end{lstlisting}
-The tree becomes
+The tree becomes:
 \begin{lstlisting}
@@
 
-The next edge is \li{(B, F, 4)}.
-The root node for \li{B} is \li{A} and the root node for \li{F} is \li{F}, so we add the edge to the tree and change the root node of \li{F} to be \li{A}.
+The next edge is \li{('B', 'F', 4)}.
+The root node for \li{'B'} is \li{'A'} and the root node for \li{'F'} is \li{'F'}, so we add the edge to the tree and change the root node of \li{'F'} to be \li{'A'}.
 
-The dictionary becomes
+The dictionary becomes:
 
@@
 \end{lstlisting}
-The next edge is \li{(E, F, 4)}.
-The root node for \li{E} is \li{C} and the root node for \li{F} is \li{A}.
+The next edge is \li{('E', 'F', 4)}.
+The root node for \li{'E'} is \li{'C'} and the root node for \li{'F'} is \li{'A'}.
 We add the edge to the tree and end the algorithm since the spanning tree now has five edges.
 
-The final spanning tree looks like
+The final spanning tree looks like:
 \begin{lstlisting}
@@
 \end{lstlisting}
-If we were to continue the algorithm, we would update the root node of \li{A} to be \li{C}. The dictionary would then look like
+If we were to continue the algorithm, we would update the root node of \li{'A'} to be \li{'C'}. The dictionary would then look like:
 
@@
 To find the root node of the tree, we look through our dictionary until we find a node that points to itself.
-For example, to find the root node of \li{F} we first need to get the value for \li{F} from the dictionary.  This is \li{A}.
-Then we get the value for \li{A} from the dictionary which is \li{C}.
-Since the value of \li{C} in the dictionary is itself, then \li{C} is the root node of the graph containing \li{F}.
+For example, to find the root node of \li{'F'} we first need to get the value for \li{'F'} from the dictionary.  This is \li{'A'}.
+Then we get the value for \li{'A'} from the dictionary which is \li{'C'}.
+Since the value of \li{'C'} in the dictionary is itself, then \li{'C'} is the root node of the graph containing \li{'F'}.
 It is necessary to trace through the dictionary in this manner each time because we avoid iterating over all of the nodes and updating their root values everytime.
@@
 
-% Note: The comments in the solutions match the pseudocode.
+% Note: The comments in the solutions match the outline.
 % When making changes, please keep that in mind.
@@
 Implement Kruskal's algorithm.
-Test your algorithm on random symmetric arrays.
-You can generate a random symmetric array by multiplying a random array by its transpose.
-Use the data from MSTdata.npy to test your tree.
-Use np.load("MSTdata.npy") to get it.
+Use the data from MSTdata.npy to test your tree. To load the data use
+\begin{lstlisting}
+np.load("MSTdata.npy")
+\end{lstlisting}
 Use the \li{formChanger} function below to put it in the right form.
@@
 
-Again, consider the example shown in Figure 7.4.
-We first initialize a dictionary with all the nodes as keys in order to track which nodes have not been processed.
-At the beginning it will be \li{\{A:False, B:False, C:False, D:False, E:False, F:False\}}.
-We then form the dictionary that maps each node to the edges that contain it.
-Since we will already know one node of each edge while we use the dictionary, we only need to store the node that is not being looked up.
-It should end up looking like \li{\{A:[(B, 3), (F, 6)], B:[(A, 3), (C, 5), (F, 4)], C:[(B, 5), (D, 1), (E, 1), (F, 5)], D:[(C, 1), (E, 2)], E:[(C, 1), (D, 2), (F, 4)], F:[(A, 6), (B, 4), (C, 5), (E, 4)]\}}.
-We will also initialize an empty dictionary to track the shortest edges that run between nodes we have processed and nodes that we haven't.
-As we iterated over the edges in our initialization step, we are also able to find the shortest edge.
-In this case, it is \li{(D, C, 1)}.
-Let's start with \li{(D, C, 1)} and initialize our tree as the list \li{[(D, C, 1)]}.
-Next, we mark \li{D} and \li{C} as processed in the dictionary that tracks which nodes have been processed.
-We now start to build our dictionary of nodes that are one edge away from our processed nodes.
-In this case, after adding the shortest edges between processed and unprocessed nodes to the dictionary, the dictionary becomes \li{\{B:(C, B, 5), E:(C, E, 1), F:(C, F, 5)\}}.
-Notice we did not include \li{E:(D, E, 2)} because there is a shorter edge to \li{E} from \li{C}.
+Again, consider the example shown in Figure \ref{mst:graph4} .
+We first initialize a dictionary with all of the nodes as keys in order to track which nodes have not been processed.
+At the beginning it will look like:
+\begin{lstlisting}
+{'A': False, 'B': False, 'C': False, 'D': False, 'E': False, 'F': False}
+\end{lstlisting}
+We set all of the values to \li{'False'} since none of the nodes have been processed yet.
 
-Of the edges in the dictionary of edges that can be processed next, the shortest is \li{(C, E, 1)}, so we add that edge to the tree and mark \li{E} as processed.
-With \li{E} being processed, we can now reach \li{F} at a cost of 4.
-After making this change, the dictionary of edges to process becomes \li{\{B:(C, B, 5), E:(C, E, 1), F:(E, F, 4)\}}.
-Since \li{E} no longer needs to be processed, we can remove it from consideration.
-So this dictionary now becomes \li{\{B:(C, B, 5), F:(E, F, 4)\}}.
+We then form another dictionary that maps each node to the edges that contain it.
+For example, the node \li{'A'} is contained in the edge \li{('A', 'B', 3)} and the edge \li{('A', 'F', 6)}, so we map \li{'A'} to these edges.
+Since we already know that \li{'A'} is in these edges, we only need to store the other nodes and weights, which in this case would be \li{('B', 3)} and \li{('F', 6)}.
+The dictionary should end up looking like:
+\begin{lstlisting}
+{'A': [('B', 3), ('F', 6)],
+ 'B': [('A', 3), ('C', 5), ('F', 4)],
+ 'C': [('B', 5), ('D', 1), ('E', 1), ('F', 5)],
+ 'D': [('C', 1), ('E', 2)],
+ 'E': [('C', 1), ('D', 2), ('F', 4)],
+ 'F': [('A', 6), ('B', 4), ('C', 5), ('E', 4)]}
+\end{lstlisting}
+We also initialize an empty dictionary to track the shortest edges that run between nodes we have processed and nodes we haven't. In total, we should now have three different dictionaries.
 
-Of the edges to be processed next, \li{(E, F, 4)} is the shortest, so we add it to the tree and mark \li{F} as processed.
-After performing the appropriate modifications to the dictionary of edges to be processed next, it becomes \li{\{B:(F, B, 4), A:(F, A, 6)\}}.
+Next we find the first edge to add to our minimum spanning tree.
+We can choose this to be the shortest edge, or edge with the smallest weight, from any given node.
+In this case, we choose \li{('D', 'C', 1)} and initialize our spanning tree as the list \li{[('D', 'C', 1)]}.
+We then mark \li{'D'} and \li{'C'} as processed in the dictionary that tracks which nodes have been processed. We do this by changing their values from \li{'False'} to \li{'True'}.
 
-Of the edges to be processed next, \li{(F, B, 4)} is the shortest, so we add it to the tree and mark \li{B} as processed.
-After performing the appropriate modifications to the dictionary of edges to be processed next, it becomes \li{\{A:(B, A, 3)\}}.
+Next we start to build our dictionary of nodes that are one edge away from our processed nodes.
+This dictionary is the one we initialized above as an empty dictionary.
+In this case, the nodes that are one edge away from \li{'D'} and \li{'C'} are \li{'B'}, \li{'F'}, and \li{'E'}.
+Notice that \li{'E'} shares an edge with both \li{'D'} and \li{'C'}.
+To deal with this, we choose the edge connected to \li{'E'} with the smallest weight, which would be the edge corresponding to \li{'C'}.
+Our dictionary then becomes:
+\begin{lstlisting}
+{'B':('C', 'B', 5), 'E':('C', 'E', 1), 'F':('C', 'F', 5)}
+\end{lstlisting}
+Again, notice that we did not include the key/value pair \li{'E':('D', 'E', 2)} because there is a shorter edge to \li{'E'} from \li{'C'}.
 
-The only edge to be considered is \li{(B, A, 3)}, so we add it to the tree.
-The tree is long enough that it it spans the nodes, so the algorithm is finished.
+Of the edges in this dictionary that can be processed next, the shortest is \li{('C', 'E', 1)}, so we add that edge to the tree and mark \li{'E'} as processed.
+With \li{'E'} being processed, we can now reach \li{'F'} at a smaller cost; four now, instead of five.
+After making these changes, the dictionary of edges to process becomes:
 
-Here's pseudocode for a version of Prim's algorithm.
-While it is not a perfectly optimized version, it is pretty good.
+\begin{lstlisting}
+{'B':('C', 'B', 5), 'E':('C', 'E', 1), 'F':('E', 'F', 4)}
+\end{lstlisting}
+
+Since \li{'E'} no longer needs to be processed, we can remove it from consideration.
+So this dictionary now becomes:
+
+\begin{lstlisting}
+{'B':('C', 'B', 5), 'F':('E', 'F', 4)}
+\end{lstlisting}
+
+Of the edges to be processed next, \li{('E', 'F', 4)} is the shortest, so we add it to the tree and mark \li{'F'} as processed.
+After performing the appropriate modifications to the dictionary of edges to be processed next, we have:
+\begin{lstlisting}
+{'B':('F', 'B', 4), 'A':('F', 'A', 6)}
+\end{lstlisting}
+
+Of the edges to be processed next, \li{('F', 'B', 4)} is the shortest, so we add it to the tree and mark \li{'B'} as processed.
+We now have:
+\begin{lstlisting}
+{'A':('B', 'A', 3)}
+\end{lstlisting}
+
+The only edge to be considered is \li{('B', 'A', 3)}, so we add it to the tree.
+The tree is long enough that it spans the nodes, so the algorithm is finished.
+
+Here's an outline for a version of Prim's algorithm.
+While it is not a perfectly optimized version, it is still effective.
 
@@
 
-\item Initialize a dictionary to track which nodes have been processed.
+\item Initialize a dictionary to track which nodes have been processed. This will look like the dictionary shown above with nodes for keys and values set to \li{'False'}.
 
-\item Initialize an empty dictionary of lists to track the edges containing each node.
+\item Create a dictionary that has for its key/value pairs all of the nodes for keys and the lists of edges each node is a part of for values. To do this first:
+	\begin{itemize}
+	\item Initialize an empty dictionary of lists.
 
-\item Fill the edge list.
-	Be sure to add each edge to the list corresponding to both of its nodes.
-
-\item Get the first edge to add (The shortest edge from any given node is a good pick).
-
+	\item Iterate through the edges of the graph to fill the lists in the above dictionary.
+		Be sure to add each edge to the lists corresponding to both of its nodes.
+	\end{itemize}
+\item Get the first edge to add. The shortest edge from any given node is a good pick.
 \item Mark the nodes in the first edge as processed.
 
-\item Initialize the tree to be the list containing the first edge.
+\item Initialize the tree to be a list containing the first edge.
 
@@
 
-	\item If that node isn't in the dictionary, set its value to be the edge passed to the functions.
+	\item If that node isn't in the dictionary, add it, and set its value to be the edge passed to the function.
 
-	\item If it is in the dictionary already, set its value to be the shorter of the edges being processed and the edges already in the dictionary.
+	\item If it is in the dictionary already, set its value to be the shorter of the two edges -  either the edge being processed currently or the edge that is already in the dictionary.
 
@@
 
-Many graph packages have functions for finding the minimal spanning tree. One of these is NetworkX. You begin as follows
+Many graph packages have functions for finding the minimum spanning tree. One of these is NetworkX. You begin as follows:
 \begin{lstlisting}
@@
 You can add a node $x$ with \li{G.add_node(x)} and add a edge from $x$ to $y$ with weight n with \li{G.add_edge(x,y,weight=n)}.
-Then \li{nx.minimum_spanning_tree(G)} returns the minimal spanning tree using Kruskals algorithm.
+Then \li{nx.minimum_spanning_tree(G)} returns the minimum spanning tree using Kruskal's algorithm.
 
 \begin{problem}
-Use NetworkX to find the minimal spanning tree of the data in MSTdata.npy.
+Use NetworkX to find the minimum spanning tree of the data in MSTdata.npy.
 Compare the output with that of your algorithm.
@@
 
-One application of Minimal Spanning Trees (MSTs) is image segmentation.
+One application of MSTs is image segmentation. Image segmentation is the process that takes an image and seperates it into similar parts.
 Kruskal's algorithm is especially good at this.
-You can convert an image into graph.
-Each pixel is a vertex and you define weights between the nodes.
-You then build the minimal spanning tree.
+To use Kruskal's we first convert the image we want to segment into a graph.
+We consider each pixel to be a node and define weights between the nodes.
+You then build the minimum spanning tree.
 If you take away the edge with the greatest weight you will get two forests (removing an edge from any graph with no cycles will create two forests).
@@
 The number of divisions often has to be higher than the actual number that is needed because sometimes one or two pixels form a division because the difference between them and the pixels around them is so great.
-You will have to adjust the number of divisions until the desired result is found.  See Figure 9.1.
+You will have to adjust the number of divisions until the desired result is found.  See Figure 5.6.
 
@@
 \caption{The original image is in the top left hand corner. The three larges segments are shown in the other corners. The original image was 498x498 and 50000 divisions were used.}
+\label{mst:graph6}
 \end{figure}

### 2014-07-27 Applications/ImageSegment/ImageSegment.tex
Small edits, mostly grammar

\lab{Applications}{Image Segmentation}{Image Segmentation}

\objective{Understand some basic applications of eigenvalues to graph theory}
\label{lab:ImgSeg_eigenvalues}

\section*{Graph Theory}
\begin{figure}[h]
\includegraphics[width=\textwidth/2]{graphExample}
\caption{A simple undirected graph}
\label{fig:example_graph}
\end{figure}

Graphs are often used to represent relationships between objects.
They are represented by a set of nodes (or vertices) and a set of edges, where each edge connects exactly two nodes.
We denote the number of vertices in a graph by $|V|$, and the number of edges by $|E|$.
A graph is \emph{directed} if connections are uni-directional, and \emph{undirected} if they are bi-directional.
One way to encode the information found in a graph is to use what is called an adjacency matrix.
\begin{definition} An adjacency matrix $A$ is a $|V| \times |V|$ matrix where the $(i,j)$-th entry $a_{ij}$ is
\begin{center}
	$a_{ij} = \begin{cases} 1 & \mbox{If an edge connects vertex i to vertex j} \\ 0 & \mbox{otherwise} \end{cases}$
\end{center}
%technically, the definition does not apply to all graphs (according to Wikipedia)
%however, this definition works if we are just considering simple graphs, i.e. at most one edge between any two vertices, no loops
\end{definition}

Every undirected edge can be represented as two directed edges.
Figure \ref{fig:example_graph} shows a simple undirected graph.
The graph in Figure \ref{fig:example_graph} can be represented as the following adjacency matrix.
\[
A = \begin{pmatrix}
0 & 1 & 0 & 0 & 1 & 0\\
1 & 0 & 1 & 0 & 1 & 0\\
0 & 1 & 0 & 1 & 0 & 0\\
0 & 0 & 1 & 0 & 1 & 1\\
1 & 1 & 0 & 1 & 0 & 0\\
0 & 0 & 0 & 1 & 0 & 0
\end{pmatrix}
\]
For undirected graphs, the adjacency matrix will always be symmetric.
The diagonal represents self-edges, or edges that connect a node to itself.

Raising the adjacency matrix to a power yields some very interesting information.
We can discover the number of paths of length $n$ between two nodes by raising a graph's adjacency matrix to the $n$th power.
For example, by squaring $A$, we can find the number of paths of length 2 between every pair of nodes.
\begin{lstlisting}
>>> A = np.array([[0,1,0,0,1,0],[1,0,1,0,1,0],
                  [0,1,0,1,0,0],[0,0,1,0,1,1],
                  [1,1,0,1,0,0],[0,0,0,1,0,0]])
>>> np.linalg.matrix_power(A,2)
array([[2, 1, 1, 1, 1, 0],
       [1, 3, 0, 2, 1, 0],
       [1, 0, 2, 0, 2, 1],
       [1, 2, 0, 3, 0, 0],
       [1, 1, 2, 0, 3, 1],
       [0, 0, 1, 0, 1, 1]])
\end{lstlisting}
We can see that no paths of length 2 exist between node 0 and node 5 because $A^2_{0,5} = 0$.
By calculating $A^6$ we can find the number of paths length 6 from node 3 to itself.
\begin{lstlisting}
>>> np.linalg.matrix_power(A, 6)
array([[45, 54, 38, 45, 54, 16],
       [54, 86, 29, 77, 51, 11],
       [38, 29, 55, 15, 70, 27],
       [45, 77, 15, 75, 31,  4],
       [54, 51, 70, 31, 93, 34],
       [16, 11, 27,  4, 34, 14]])
\end{lstlisting}
We see that there are 55 unique paths of length 6 from node 3 to itself.
Imagine trying to count all of those paths by hand!
It would be very easy to count incorrectly.
However, this method makes it very simple to count paths without any mistakes.

Adjacency matrices can also be composed of \li{True} and \li{False} values.
In this case, the $n$th power of such a matrix (using boolean arithmetic)
is again a matrix of
boolean values which simply indicate whether there exists a path of length $n$ between the given pair of nodes, rather than indicating the number of such
paths.

\begin{problem}
Let the following matrix represent a directed graph
\[
\begin{pmatrix}
0 & 0 & 1 & 0 & 1 & 0 & 1 \\
1 & 0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 & 0 \\
1 & 0 & 0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 & 0 \\
0 & 0 & 1 & 0 & 0 & 0 & 1 \\
0 & 1 & 0 & 0 & 0 & 0 & 0
\end{pmatrix}
\]
Between which pair of nodes does there exist the greatest number of paths
of length five?
From which node to which node is there no path of length seven?
\end{problem}

Another useful way of representing an undirected graph is a Laplacian matrix.
This special matrix can reveal a lot of information about a graph.
\begin{definition}
For a simple graph (an unweighted, undirected graph without self-edges), the Laplacian of a graph $G$ is
\[ L_G = D_G - A_G \]
where $D_G$ is the degree matrix of $G$ and $A_G$ is the adjacency matrix of $G$.
\end{definition}
An important question in working with graphs is the degree of its nodes.
The degree of a node is the number of edges that connect to the node.
For a directed graph, each node has an \emph{out-degree} (the number of edges directed away from a node) and an \emph{in-degree} (the number edges directed toward a node).
The degree matrix of a graph is a diagonal matrix that has the degree of each node as the diagonal entries (thus, for a directed graph there in an in-degree matrix and an out-degree matrix).

\begin{problem}
Calculate the Laplacian matrix of the graph in Figure \ref{fig:example_graph} by calculating $D$ and $A$.
Think about how to find the number of neighboring nodes using an adjacency matrix.

Write a general solution that will calculate the Laplacian of any small graph.
Use the adjacency matrix to check that the graph is unweighted, undirected, and contains no self-edges.
\label{prob:laplacian}
\end{problem}

We may also define weighted graphs so that each edge has a weight or cost attached to it.
In this case the adjacency matrix will have the cost in each entry, rather than just a 1 or 0.
An example of a weighted may be a collection of cities as vertices, roads connecting them as edges, and the distance of the connecting roads as the cost attached to each edge.

In this lab we will be investigating the sorts of things that we can discover about a graph by examining the spectrum of its adjacency matrix.
While the formulation of the adjacency matrix seems to be very simple, its structure allows us to learn surprising things by examining its eigenvalues.

For a weighted graph the definition of the Laplacian of a graph is:

\begin{definition}  The \emph{degree} of a vertex of a weighted graph is the sum of the weights of the edges connected to that vertex.
In undirected graphs, which is what we will be dealing with, this is the sum of all the weights moving into that vertex.
This definition varies slightly for directed graphs.
Let $D$ be a diagonal matrix with
\[
D_{ii} = \mbox{ Degree of vertex $i$}
\]
and let $A$ be the adjacency matrix of the graph.
The graph \emph{Laplacian} $Q$ is given by
\[
Q = D-A
\]
\end{definition}

The Laplacian matrix of a graph will typically be very sparse.
We will be examining the spectrum of the Laplacian matrix.

A connected graph is a graph where every vertex is connected to every other vertex by at least one path.
An important question one might ask about a graph is whether or not it is connected.
What is the best way to do this?
A naive approach would be to exhaustively map every possible path from each vertex.
While this would be feasible for very small graphs, most interesting graphs (for example, the internet) will have thousands of vertices and such an approach becomes essentially impossible to execute.

It turns out there is a better way.
If the second smallest eigenvalue of the Laplacian matrix associated with a graph is positive, then this graph is connected.
The mathematics behind this is quite involved, so we omit a proof for now.
In many applications the Laplacian matrix will be very sparse, and thus with optimizations made for sparsity, we can discover the connectivity of a graph relatively cheaply.

\begin{problem}Write a function \li{laplacian} that accepts an adjacency matrix as an argument and returns the Laplacian matrix and its second smallest eigenvalue.
Use the \li{scipy.linalg} package to compute this eigenvalue.
Using {\tt numpy.rand(n,n)}, generate several random matrices of various sizes.
Use masking to change the sparsity of said matrices, i.e. {\tt numpy.rand(n,n) > c} for {\tt c = 0.25,0.5,0.75}.
Try different distributions and $c$ values.
What can we expect about the connectivity of random matrices?
\end{problem}

A related problem is found in image processing.
An image is a collection of coordinates and light intensities.
We call each coordinate and associated brightness a pixel.
We let every pixel in an image be a vertex in a graph that is connected to its neighbors within a certain radius.
For an $N \times N$ image, we define an adjacency matrix as follows:

\begin{equation}
\label{eq:adjacency}
w_{ij} = e^{-\frac{|I(i) - I(j)|}{\sigma_I^2}} \cdot \begin{cases} e^{-\frac{d(i,j)}{\sigma_d^2}} & \mbox{ for $d(i,j) < r$} \\ 0 & \mbox{ otherwise} \end{cases}
\end{equation}
for $i = 1 \hdots N^2$ and $j = 1 \hdots N^2$ (traveling through each pixel in the image), where
\begin{itemize}
	\item$d(i,j)$ is the Euclidean distance between pixel $i$ and pixel $j$
	\item $|I(i) - I(j)|$ is the difference in brightness of pixels $i$ and $j$
	\item $\sigma_I$ and $\sigma_d$ are constants
\end{itemize}
Thus, given a $N\times N$ image, we will produce an adjacency matrix of size $N^2\times N^2$.
Even for smaller images, this will become very large, but we have sparsity on our side.
Even so, only do this for images smaller than 50 by 50.
Since each pixel is only connected to the pixels within radius $r$ around it, relatively few entries will be non-zero.

\begin{problem}
Write a function \li{adjacency} that takes an $N \times N$ image array, radius $r$, and values for
$\sigma_I^2$ and $\sigma_d^2$, and returns the adjacency matrix defined in \eqref{eq:adjacency}.
Notice that for each pixel you can save time by only checking the pixels $r$ rows and columns away.
%For that you'll have to handle the pixels on the edges and corners of the image carefully.
%I gave them new helper code, which abstracts away the edge cases.
Make use of sparsity in order to feasibly store such a large matrix.
We have provided code to help out with this problem.
\end{problem}

An important problem in image processing is that of image segmentation.
When humans observe an image, they can easily pick out portions of an image that ``belong together.''
For example, if we saw the picture of a person against a black background, the pixels making up the background would make up one segment of an image, and the person would make up the other part.
While this seems simple for a human, how can we program a computer to do so?

Once again, we turn to the eigenvalues of the Laplacian matrix.
We examine the second smallest eigenvalue of $D^{-\frac{1}{2}}QD^{-\frac{1}{2}}$, where $D$ is the degree matrix and $Q$ is the Laplacian matrix of the adjacency matrix defined in \eqref{eq:adjacency}.
The associated $N^2 \times 1$ eigenvector will have positive and negative entries, splitting the image into two parts.
Thanks to some very fancy mathematics and the way we have defined the weights between nodes, this does a pretty good job of breaking an image into two segments.
Recursively executing this function, we can find segments within segments.
%Figure \ref{segmentation:example} shows a simple example of an image and its segments.

%\begin{figure}
%\includegraphics[scale=0.2]{monument}
%\includegraphics[scale=0.2]{segment1}
%\includegraphics[scale=0.2]{segment2}
%\caption{An image and it's segments.}
%\label{segmentation:example}
%\end{figure}

\begin{problem}  Write a function \li{segment} that solves the segmentation problem for small images.
Accept an image array as an argument and return the two segments.
Use $r = 5, \sigma_I^2 = 0.02,$ and $\sigma_d^2 = 3.0$.
Remember that the Laplacian matrix will be very large but also very sparse.
Because of the way we defined $\sigma_I^2$ the image matrix intensity values need to be between $0.0$ and $1.0$.
\end{problem}

### 2014-07-29 Algorithms/EigSolver/Eig.tex
updating file

\lab{Algorithms}{Eigenvalue Solvers And Markov}{Eigenvalue Solvers And Markov}
\objective{Implement the $QR$ algorithm for finding eigenvalues and learn Markov chains.}
\label{lab:EigSolve}

\section*{Eigenvalues are hard to find}

Finding the eigenvalues of $n \times n$ matrix $A$ means solving the following equation, where $x$ is a nonzero vector and $\lambda$ is a scalar.
\begin{align}
A x &= \lambda x \notag \\
A x - \lambda x &= 0 \notag \\
(A - \lambda I)x  &= 0 \label{eq:singularity}
\end{align}
Since $x$ is nonzero, \eqref{eq:singularity} indicates $A-\lambda I$ must be singular.
Thus, $\det(A - \lambda I) = 0$.  This determinant is often notated $\det(A - \lambda I) = p(\lambda)$ and is called the \emph{characteristic polynomial} of $A$.
The roots of the characteristic polynomial are the eigenvalues of $A$.

If $A$ is $n \times n$, the degree of $p(\lambda)$ is  $n$.
Finding the roots is easy for small $n$, but it becomes difficult or impossible as $n$ increases.
Abel's Theorem  outlines the problem.

\begin{theorem}
{\bf Abel's Impossibility Theorem:} There is no general algebraic solution for solving a polynomial equation of degree $n>4$.
\label{Theorem:Abel}
\end{theorem}

Therefore, there is no method that will exactly find the eigenvalues of an arbitrary matrix.
This is a significant result. In practice, it means that we often rely on iterative methods, which converge to the eigenvalues.

\section*{The Power Method}
There are many such iterative methods for finding eigenvalues. The Power Method is a simple, fast way to find the largest eigenvalue and eigenvector. Beginning with a vector $x_0$ that has norm 1, the following iteration is applied:
\[
x_{k+1}=\frac{Ax_k}{\norm{Ax_k}}
\]
it works when
\begin{itemize}
\item The matrix $A$ has a eigenvalue that is strictly greater in magnitude than its other eigenvectors.
\item your starting vector $x_0$ has a nonzero component in the direction of the eigenvector of the dominant eigenvalue.

\end{itemize}

This often converges slowly (geometric with the ration given by $\norm{\lambda_2/\lambda_1}$, but is especially useful when your matrix has all positive values because you are guaranteed to have a dominant eigenvalue. Note that the modulus of the dominant eigenvalue is often much greater than the rest of the eigenvalues so the convergence is much faster.
Given an eigenvector, its corresponding eigenvalue can be computed using the Rayleigh quotient:
\[
\lambda = \frac{Ax\cdot x}{x\cdot x}
\]
\begin{problem}
Write a method that takes a in matrix and a tolerance and computes its largest eigenvalue and the eigenvector. Have your starting vector be random. Use the two-norm and have the iteration stop when the maximun change of the vector is less than the tolerance. Test your function on positive matrices.
\end{problem}

\begin{comment}
An overview of the proof of the method is that you can write a matrix in Jordan Conical form $A=VJV^{-1}$ where $V$ is the matrix of the generalized eigenspaces. But the first column is is the eigenvector corresponding to largest eigenvalue and $J$ is a upper trianglar matrix of eigenvalues and ones.  Note that $A^k=VJ^kV^{-1}$. The limit as $k \rightarrow \infty$ of $(\frac{1}{\lambda_1}J)^k$ is a matrix of all zeros except for a one in the upper right hand corner. So $(\frac{A}{\norm{A}})^k \approx VJ^kV^{-1}$ So the largest eigenvalue dominates.
\end{comment}

The down side is that it only finds the largest eigenvector and eigenvalue. The next algorithm finds more eigenvalues.

\section*{The $QR$ algorithm}

We will explore one of the simplest: the $QR$ algorithm.
The following recurrence describes the $QR$ Algorithm in its most basic form:
\begin{equation*}
A_0 = A, \quad A_k = Q_k R_k, \quad A_{k+1} = R_k Q_k
\end{equation*}
where $Q_k, R_k$ is the $QR$ decomposition of $A_k$.
Yes, it's as easy as it looks.
All this algorithm does at each step is find the $QR$ decomposition of $A_k$ and multiply $Q_k$ and $R_k$ together again but in the opposite order.
How does this simple algorithm find the eigenvalues of $A$?

\begin{comment}
\begin{problem}
\label{problem:similarity proof}
Prove that $A_{k+1} \sim A_k$ (where $\sim$ denotes matrix similarity).
Then prove that $A_n \sim A$ for all $n$.
\end{problem}
\end{comment}

Observe that $A_{k+1} \sim A_k$ (where $\sim$ denotes matrix similarity).
Then $A_n \sim A$ for all $n$.
This statement shows that $A_n$ has the same eigenvalues as $A$.
Preservation of eigenvalues is the first important feature that makes the algorithm work.
The other important feature is that each iteration of the algorithm effectively transfers some of the``mass" from the lower to the upper triangle.
Under very general conditions, $A_n$ will converge to a matrix of the form

\begin{equation*}
\label{eq:Schur form}
S =
     \begin{pmatrix}
          S_1 &* & \cdots & * \\
           0     &S_2  &  \ddots & \vdots \\
           \vdots  & \ddots & \ddots & *  \\
           0 & \cdots & 0 & S_m
    \end{pmatrix}
\end{equation*}
where $S_i$ is either a $1 \times 1$ or a $2 \times 2$ matrix.
For most matrices $A$, all the $S_i$ will be $1 \times 1$, so $S$ will be an upper triangular matrix.
In this case, $S$ is called the \emph{Schur form} of $A$.
The eigenvalues of $A$ are on the main diagonal of $S$.

The only case where $S$ is not upper triangular is when $A$ is a real but not symmetric matrix.
In this case, though $A$ is real, it may have complex eigenvalues.
These eigenvalues occur in complex conjugate pairs.
Each of these pairs corresponds to a $2 \times 2$ block in $S$, where the eigenvalues of the $2 \times 2$ block are the complex conjugate pair of eigenvalues of $A$.
In this case, $S$ is called the \emph{real Schur form} of $A$.

\subsection*{Hessenberg preconditioning}

Recall that an upper Hessenberg matrix looks like
\[
\begin{pmatrix}
* & * & * & \cdots & * \\
* & * & * & \cdots & * \\
0 & * & * & \cdots&* \\
\vdots & & \ddots & \ddots & \vdots \\
0 & \cdots & 0 & * & *\\
\end{pmatrix}
\]
and that every matrix is similar to an upper Hessenberg matrix.
Hessenberg reduction also preserves eigenvalues.
It is a good idea to reduce to Hessenberg form before continuing with the $QR$ algorithm.
You'll converge to the Schur form faster this way, since Hessenberg matrices are already close to upper triangular.

\begin{problem}
Write a function \li{qrSolver} that implements the QR algorithm as described above. Use the QR decomposition built into \li{scipy.linalg}.
Have your function accept a real-valued $n \times n$ matrix $A$, a number of iterations, and a tolerance number,
and return the corresponding estimate for all the eigenvalues of $A$.
Note that you will need to find the eigenvalues of the $2 \times 2$ $S_i$ directly, whenever there are any.
\end{problem}

Reducing the matrix to Schur form also has the added benefit that each iteration for an $n \times n$ array can be computed in $\mathcal{O} \left( n^2 \right)$ time instead of $\mathcal{O} \left( n^3 \right)$.
One algorithm for computing the QR decomposition of an upper Hessenberg matrix was discussed in Lab \ref{lab:givens}.
This works since $Q$ in the QR factorization of an upper Hessenberg matrix is also upper Hessenberg, the product $R Q$ also turns out to be upper Hessenberg.

\begin{comment}
\begin{problem}
\label{prob:QR_eig_hessenberg}
Write a version of the QR algorithm that performs the QR algorithm by computing the Hessenberg form of a matrix, then computing various QR decompositions of the Hessenberg form of the matrix.
Use your solutions to \ref{prob:hessenberg} (where you computed the Hessenberg form of a matrix) and Problem \ref{prob:givens_hessenberg_modified} to do the necessary computations (where you computed the QR decomposition of a Hessenberg matrix and wrote code for multiplication by $Q$ that works in $\mathcal{O} \left( n^2 \right)$ time).
The solution to Problem \ref{prob:givens_hessenberg_modified} is especially important because it allows the compution of each QR decomposition and each $R Q = \left( Q^T R^T \right)$ in $\mathcal{O} \left( n^2 \right)$ time.
\end{problem}
\end{comment}

\begin{comment}

\begin{problem}
If $A$ is normal, its Schur form is diagonal.
For normal $A$, have your function additionally output the eigenvector corresponding to each eigenvalue.
Hint 1: Test your function on Hermitian and real symmetric matrices; they are both normal.
Hint 2: Your work in Problem \ref{problem:similarity proof} will help.
You have already made all the necessary calculations, you just need to store the information correctly.
\end{problem}

\end{comment}

\begin{problem}
Test your implementation with random matrices.
Try real-valued and symmetric matrices.
Compare your output to the output from the eigenvalue solver.
How many iterations are necessary?
How large can $A$ be?
\end{problem}

Be aware that the algorithm we coded today is a very un-optimized approach.
In practice, the QR algorithm as we have described it here is not generally used.
There is an improved version called the Implicit QR algorithm that iterates much more efficiently.

Further, the QR algorithm is not the only iterative method used to find eigenvalues.
Arnoldi iteration is similar to the QR algorithm but exploits sparsity.
Other methods include the Jacobi method and the Rayleigh quotient method.

As a final note, it is important to remember that eigenvalue solvers can be wrong, particularly for matrices that are ill-conditioned. Convergence is not unconditionally guaranteed.

\section*{Markov Chains}
A Markov Chain describes a particular type of random process that
undergoes a sequence of transitions among various states. It is
characterized by the fact that all relevant information is related to its current state.
We can easily model this process using matrices.
We will start with a simple example of a frog jumping from one lily pad to another.

Suppose Fredo the frog jumps around between the three lily pads 1, 2, and 3.
These three pads are taken as the possible \emph{states} of the system, and the
lily pad on which Fredo is presently sitting is the \emph{current state}.
If Fredo is on lily pad 1 and jumps, there is a 25\% chance that it will land back on lily pad 1, a 25\% chance that it will land on lily pad 2, and a 50\% chance that it will land on lily pad 3.
We can find similar probabilities if he starts on lily pad 2 or 3.
Such probabilities are known as the \emph{transition probabilities}.
In figure \ref{fig:markov1} we have a transition diagram that reflects the various probabilities of jumping from one lily pad to another.

\begin{figure}[h]
\includegraphics[width=\textwidth]{markov1}
\caption{Transition diagram for Fredo the Frog}
\label{fig:markov1}
\end{figure}

We can convert our transition diagram into a transition matrix, where the $(i,j)$-entry of the matrix corresponds to the probability that Fredo jumps from the $j^{th}$ lily pad to the $i^{th}$ lily pad (where $1$ is the first lily pad, $2$ is the second, and so on).
The transition matrix is
\[
A = \begin{pmatrix}
1/4 & 1/2 & 1/2\\
1/4 & 1/6 & 1/2\\
1/2 & 1/3 & 0
\end{pmatrix}
\]
Note that all of the columns add up to one.
This is important.

If Fredo is on lily pad 1, where will it be after two jumps?
By multiplying the matrix $A$ by itself, we have (approximately)

\[
A^2 = \begin{pmatrix}
0.4375 & 0.3750 & 0.3750\\
0.3542 & 0.3194 & 0.2083\\
0.2083 & 0.3056 & 0.4167
\end{pmatrix}
\]
From this, we infer that there is a 43.75\% chance the frog will still be on lily pad 1 after two jumps.
Note that it might have jumped from 1 to 1 to 1, denoted $1 \rightarrow 1 \rightarrow 1$, or it could have jumped to one of the other lily pads and then back again, that is, either $1 \rightarrow 2 \rightarrow 1$ or $1 \rightarrow 3 \rightarrow 1$.
In addition, there is a 35.42\% chance it will be on lily pad 2 and a 20.83\% chance that it will be on lily pad 3.
Using Python, we can type in our transition matrix and see where Fredo will be after 5, 10, 20 or 100 jumps.

\begin{lstlisting}
#Remember, the 1.'s in the numerator force floating point division
>>> A = np.array([[1./4,1./2,1./2],[1./4,1./6,1./2],[1./2,1./3,0]])
>>> np.linalg.matrix_power(A,5)
>>> np.linalg.matrix_power(A,10)
>>> np.linalg.matrix_power(A,20)
>>> np.linalg.matrix_power(A,100)
\end{lstlisting}

Note that as we take higher powers it appears that the limit goes to
\[
A^\infty = \begin{pmatrix}
0.4 & 0.4 & 0.4\\
0.3 & 0.3 & 0.3\\
0.3 & 0.3 & 0.3
\end{pmatrix},
\]
and this can, in fact, be proven carefully.
This means that after several jumps, the probability that we will find Fredo on a given lily pad will have nothing to do with where he started initially.

We can generalize this notion beyond that of frogs and lily pads.
Let the state distribution of our system be represented by a probability vector
\[
\x = \begin{bmatrix}
x_1\\
x_2\\
\vdots\\
x_n
\end{bmatrix}
\]
where each entry represents the probability of being in that state.
Note that each entry is non-negative and the sum of all the entries adds up to one.
For example, in the frog example, if we know initially that it is on lily pad 1, then we have the probability vector
\[
\x_0 = \begin{bmatrix}
1\\
0\\
0
\end{bmatrix}
\]
because we know for certainty (100\%) that Fredo is in the first state.
After one jump, we have
\[
\x_1 = A \x_0 = \begin{bmatrix}
0.25\\
0.25\\
0.50
\end{bmatrix}
\]
After two jumps, we have
\[
\x_2 = A \x_1 = A^2 \x_0 = \begin{bmatrix}
0.4375\\
0.3542\\
0.2083
\end{bmatrix}
\]
After a large number of jumps $(n>>1)$, we have
\[
\x_n = A \x_{n-1} = \dots = A^n \x_0 \approx \begin{bmatrix}
0.4\\
0.3\\
0.3
\end{bmatrix}
\]
Since all of the columns are the same for $A^\infty$, it follows that for any initial probability vector $\x_0$, we get the same limiting output, or in other words, all initial vectors converge to the same point, call it $\x_\infty$.
Moreover, we have that
\[
\x_\infty = A \x_\infty
\]
This is called a \emph{stable fixed point}.
How can we check that a stable fixed point exists?

Notice this is just the power method iteration without dividing by the norm. Since the all of the columns of $A$ sum to one and $x_0$ sums to one the one norm of $Ax_k=1$ for all $k$. So this goes to a stable fixed point and that fixed point turns out to be the eigenvector corresponding to the largest eigenvalue. Note that all the columns sum up to one and that a matrix and its transpose have the same eigenvalues. It can be shown that if all the rows of a positive matrix sum to the same number that number is the largest eigenvalue and the corresponding eigenvector is the one vector. So the $A.T$ has the eigenvalue $\lambda=1$ so $A$. So for any Markov chain the largest eigenvalue will be $\lambda=1$.

\section*{Example}
Consider the Markov chain given by
\[
A = \begin{pmatrix}
0.5 & 0.3 & 0.4\\
0.2 & 0.2 & 0.3\\
0.3 & 0.5 & 0.3
\end{pmatrix}.
\]

We do this via Python:
\begin{lstlisting}
>>> from scipy import linalg as la
>>> A = np.array([[.5,.3,.4],[.2,.2,.3],[.3,.5,.3]])
>>> V = la.eig(A)[1]
\end{lstlisting}
Note that the entries in the $\lambda=1$ eigenvector do not generally add up to one.
Indeed, any multiple of an eigenvector is an eigenvector.
So we need to multiply it by the appropriate constant so that all of the entries add up to one.
\begin{lstlisting}
>>> x = V[:,0]
>>> x = x/np.sum(x);x
array([ 0.41836735,  0.23469388,  0.34693878])
\end{lstlisting}
We can check this answer by taking $A$ to a high exponent, say $A^{100}$.

\begin{problem}
Suppose a basketball player's success at shooting free throws can be described with the following Markov chain
\[
A = \begin{pmatrix}.75&.50\\.25&.50\end{pmatrix}
\]
where the first state corresponds to success and the second state to failure.
\begin{enumerate}
\item If the player makes his first free throw, what is the probability that he also makes his third one?
\item What is the player's average free throw percentage?
\end{enumerate}
\end{problem}

\begin{problem}
Consider the Markov process given by the transition diagram in Figure \ref{fig:markov2}.
\begin{figure}[H]
\includegraphics[width=\textwidth]{markov2}
\caption{Transition diagram}
\label{fig:markov2}
\end{figure}

\begin{enumerate}
\item Find the transition matrix.
\item If the Markov process is in state 1 initially, find the probability that it is in state 2 after 2 transitions.
\item Find the stable fixed point if it exists.
\end{enumerate}
\end{problem}

### 2014-08-03 Applications/ImageSegment/ImageSegment.tex
Small edits, to explain some concepts better

@@
 Figure \ref{fig:example_graph} shows a simple undirected graph.
-The graph in Figure \ref{fig:example_graph} can be represented as the following adjacency matrix.
+The graph in Figure \ref{fig:example_graph} can be represented by the following adjacency matrix.
 \[
@@
 We can discover the number of paths of length $n$ between two nodes by raising a graph's adjacency matrix to the $n$th power.
-For example, by squaring $A$, we can find the number of paths of length 2 between every pair of nodes.
+For example, by squaring $A$, we can find the number of paths of length two between every pair of nodes.
 \begin{lstlisting}
@@
                   [1,1,0,1,0,0],[0,0,0,1,0,0]])
+
 >>> np.linalg.matrix_power(A,2)
@@
 \end{lstlisting}
-We can see that no paths of length 2 exist between node 0 and node 5 because $A^2_{0,5} = 0$.
-By calculating $A^6$ we can find the number of paths length 6 from node 3 to itself.
+We can see that no paths of length two exist between node 0 and node 5 because $A^2_{0,5} = 0$.
+By calculating $A^6$ we can find the number of paths of length six from node 3 to itself.
 \begin{lstlisting}
@@
 \end{lstlisting}
-We see that there are 55 unique paths of length 6 from node 3 to itself.
+We see that there are 75 unique paths of length six from node 3 to itself.
 Imagine trying to count all of those paths by hand!
 It would be very easy to count incorrectly.
-However, this method makes it very simple to count paths without any mistakes.
+This method makes it very simple to count paths without mistakes.
 
@@
 
+An important question in working with graphs is the degree of its nodes.
+The degree of a node is the number of edges that connect to the node.
+For a directed graph, each node has an \emph{out-degree} (the number of edges directed away from a node) and an \emph{in-degree} (the number edges directed toward a node).
+The degree matrix of a graph is a diagonal matrix that has the degree of each node as the diagonal entries (thus, for a directed graph there is an in-degree matrix and an out-degree matrix).
+
 Another useful way of representing an undirected graph is a Laplacian matrix.
@@
 \begin{definition}
-For a simple graph (an unweighted, undirected graph without self-edges), the Laplacian of a graph $G$ is
+For a simple graph (an undirected graph without self-edges), the Laplacian of a graph $G$ is
 \[ L_G = D_G - A_G \]
@@
 \end{definition}
-An important question in working with graphs is the degree of its nodes.
-The degree of a node is the number of edges that connect to the node.
-For a directed graph, each node has an \emph{out-degree} (the number of edges directed away from a node) and an \emph{in-degree} (the number edges directed toward a node).
-The degree matrix of a graph is a diagonal matrix that has the degree of each node as the diagonal entries (thus, for a directed graph there in an in-degree matrix and an out-degree matrix).
 

### 2014-08-03 Algorithms/MST/mst.tex
small grammatical errors
Clarified the beginning of image segmentation and improved instructions on problem 1

@@
 
-One application of MSTs is image segmentation. Image segmentation is the process that takes an image and seperates it into similar parts.
-Kruskal's algorithm is especially good at this.
-To use Kruskal's we first convert the image we want to segment into a graph.
-We consider each pixel to be a node and define weights between the nodes.
+One application of MSTs is image segmentation.
+Image segmentation is the process that takes an image and seperates it into similar parts.
+This process has many applications.
+For example, when analyzing a satellite image, it might be useful to distinguish between the different parts of the image to aid in object recognition.
+
+Kruskal's algorithm is especially good at segmenting images.
+To use Kruskal's, we first need to convert the image we want to segment into a graph.
+We do this by considering each pixel to be a node.
+To find the weights between each node, we take the absolute value of the difference between the values of the pixels.
+For example, consider an image represented by the following matrix:
+
+\[
+\begin{pmatrix}
+37 & 44 & 40 & 39 \\
+50 & 51 & 45 & 45 \\
+55 & 45 & 45 & 54 \\
+61 & 59 & 54 & 52 \\
+\end{pmatrix}
+\]
+
+We first look at the nodes 0 and 1, corresponding to the pixel values 37 and 44.
+Our first edge would then be \li{(0, 1, 7)}, where 7 is the absolute value of the difference between 37 and 44.
+We then continue on the first row, finding the weights between nodes.
+Once we reach the second row we will be on node 4.
+The edge between nodes 4 and 5 will then be the edge between pixel values 50 and 51 and will therefore have a corresponding weight of 1.
+We continue in this manner, until we convert every pixel into a node and find the weight between nodes, or in this case, the difference in pixel values.
+
+There are many different ways to turn an image into a graph and weight the edges.
+This version guarantees however, that there will be less than $4n$ edges, where $n$ is the number of nodes.
+Other image segmentation algorithms have to use $n^2$ space.
+This gives the MST algorithm a critical advantage over other image segmentation algorithms.
+
+\begin{problem}
+Write a function that takes a black and white image as input and outputs a list of the edges and corresponding weights. Store the edges using the form \li{(node,node,weight)}.
+
+Apply your function to the image saved as MSTseg1.jpg.
+This image is read in as a three-dimensional array, but we want a two-dimensional array in order to convert the image into a graph.
+To fix this we average over the third axis to reduce the dimensions.
+\begin{lstlisting}
+from matplotlib import pyplot as plt
+img = plt.imread('MSTseg1.jpg')
+img2 = img.mean(axis=2)
+\end{lstlisting}
+
+\begin{figure}[H]
+\includegraphics[width = .4\textwidth]{MSTseg1.jpg}
+\caption{MSTseg1.jpg}
+\end{figure}
+
+\end{problem}
+
 You then build the minimum spanning tree.
@@
 Kruskal's algorithm is performed until $n-(k+1)$ edges are added which is the same as taking out the $k$ edges with greatest weights.
-
-There are many different ways to turn an image into a graph and weight the edges.
-A simple, yet effective, version is to make every pixel a node and the edges are the difference in intensities in the four cardinal directions.
-
-This means that there are less than $4n$ edges.
-Other image segmentation algorithms have to use $n^2$ space.
-This gives the MST algorithm a critical advantage over other image segmentation algorithms.
-
-\begin{problem}
-Write a function that takes a black and white image as input and outputs a list of the edges and a list of nodes.
-Store the edges using the form \li{(node,node,weight)}.
-\end{problem}
 

### 2014-08-09 Algorithms/MST/mst.tex
Finished up MST Image Segmentation. Added two files as helper functions. Clarified a couple of parts including problems 4 and 5 and deleted one problem

@@
 \caption{An example of an undirected graph}
-\label{mst:graph1}
+\label{fig:graph1}
 \end{figure}
@@
 A graph is \emph{undirected} if its edges are bi-directional. This means that the edges have no specific direction in between nodes. Figure
-\ref{mst:graph1} shows an example of an undirected graph.
+\ref{fig:graph1} shows an example of an undirected graph.
 A \emph{weighted graph} is a graph where each edge has a value associated with it.
@@
 If there is no edge, we put a 0.
-For the above graph in Figure \ref{mst:graph1} we generate the following matrix:
+For the above graph in Figure \ref{fig:graph1} we generate the following matrix:
 
@@
 
-Now consider the graph in Figure \ref{mst:graph4}.  This graph is similar to the graph in Figure \ref{mst:graph1}, except now weights are attached to each edge.  
+Now consider the graph in Figure \ref{fig:graph2}.  This graph is similar to the graph in Figure \ref{fig:graph1}, except now weights are attached to each edge.  
 \begin{figure}[H]
@@
 \caption{A weighted, undirected graph}
-\label{mst:graph4}
+\label{fig:graph2}
 \end{figure}
@@
 For a weighted, undirected graph, a third value could be added to each pair of nodes to represent the corresponding weight of the edge.
-A list like this for the graph in Figure \ref{mst:graph1} would look like this:
+A list like this for the graph in Figure \ref{fig:graph1} would look like this:
 
@@
 A cycle, for undirected graphs, is a path that starts and ends on the same node without crossing any edge more than once.
-Figure \ref{mst:graph3} shows an example of an undirected graph with a cycle in it.
+Figure \ref{fig:graph3} shows an example of an undirected graph with a cycle in it.
 The cycle includes the nodes A, B, C, and F.
@@
 \caption{A cycle in an undirected graph}
-\label{mst:graph3}
+\label{fig:graph3}
 \end{figure}
@@
 Both Kruskal's and Prim's Algorithms are methods that find a minimum spanning tree of a weighted, undirected graph.
-Figure \ref{mst:graph2} shows a spanning tree of the graph shown in Figure \ref{mst:graph1}.
-Figure 5.5 shows a minimum spanning tree of the graph shown in Figure \ref{mst:graph4}.
+Figure \ref{fig:graph4} shows a spanning tree of the graph shown in Figure \ref{fig:graph1}.
+Figure \ref{fig:graph5} shows a minimum spanning tree of the graph shown in Figure \ref{fig:graph2}.
 
@@
 \includegraphics[width = .4\textwidth]{graph2.pdf}
-\caption{A spanning tree for the graph in Figure \ref{mst:graph1}.}
-\label{mst:graph2}
+\caption{A spanning tree for the graph in Figure \ref{fig:graph1}.}
+\label{fig:graph4}
 \end{figure}
@@
 \includegraphics[width = .4\textwidth]{graph5.pdf}
-\caption{A MST of the graph in Figure \ref{mst:graph4}.}
-\label{mst:graph5}
+\caption{A MST of the graph in Figure \ref{fig:graph2}.}
+\label{fig:graph5}
 \end{figure}
@@
 
-We will consider the graph in Figure \ref{mst:graph4} and apply Kruskal's algorithm.
+We will consider the graph in Figure \ref{fig:graph2} and apply Kruskal's algorithm.
 First, we initialize our spanning tree to be an empty list. Then, to track the root nodes of each connected tree, we initialize a dictionary where each node points to itself.
@@
 
-Again, consider the example shown in Figure \ref{mst:graph4} .
+Again, consider the example shown in Figure \ref{fig:graph2}.
 We first initialize a dictionary with all of the nodes as keys in order to track which nodes have not been processed.
@@
 To find the weights between each node, we take the absolute value of the difference between the values of the pixels.
-For example, consider an image represented by the following matrix:
+For example, consider an image represented by the following matrix, where each element is a pixel:
 
@@
 \]
-
-We first look at the nodes 0 and 1, corresponding to the pixel values 37 and 44.
-Our first edge would then be \li{(0, 1, 7)}, where 7 is the absolute value of the difference between 37 and 44.
-We then continue on the first row, finding the weights between nodes.
-Once we reach the second row we will be on node 4.
-The edge between nodes 4 and 5 will then be the edge between pixel values 50 and 51 and will therefore have a corresponding weight of 1.
-We continue in this manner, until we convert every pixel into a node and find the weight between nodes, or in this case, the difference in pixel values.
+To number our nodes, we first assign node 0 to the pixel in the upper left corner of the matrix and move across the row until we have assigned each pixel a node, ending with node 3.
+We then start at the begninning of the second row and assign node 4 to the first pixel in that row and continue on in this manner until all of the pixels in the matrix have been assigned a node.
+To find the weights between each node we start by calculating the weight between nodes 0 and 1, corresponding to the pixel values 37 and 44.
+Our first edge would then be \li{(0, 1, 7)}, where 7 is the weight, or the absolute value of the difference between 37 and 44.
+We then find the weight between nodes 0 and 4, which we calculate to be 13.
+Our second edge therefore would be \li{(0, 4, 13)}.
+We continue on in this manner, until we have found all of our edges.
 
@@
 This version guarantees however, that there will be less than $4n$ edges, where $n$ is the number of nodes.
-Other image segmentation algorithms have to use $n^2$ space.
+Other image segmentation algorithms are forced to use $n^2$ space.
 This gives the MST algorithm a critical advantage over other image segmentation algorithms.
@@
 \begin{problem}
-Write a function that takes a black and white image as input and outputs a list of the edges and corresponding weights. Store the edges using the form \li{(node,node,weight)}.
+Write a function that takes a black and white image as input and outputs a list of the edges and corresponding weights. Store the edges and weights using the form \li{(node,node,weight)}.
 
-Apply your function to the image saved as MSTseg1.jpg.
-This image is read in as a three-dimensional array, but we want a two-dimensional array in order to convert the image into a graph.
-To fix this we average over the third axis to reduce the dimensions.
+Apply your function to the image saved as `MSTseg1.jpg' (the top left image in Figure \ref{fig:graph6}).
+This image is read in as a three-dimensional array.
+Since we want a two-dimensional array in order to convert the image into a graph, we average over the third axis to reduce the dimensions.
+
 \begin{lstlisting}
@@
 \end{lstlisting}
-
-\begin{figure}[H]
-\includegraphics[width = .4\textwidth]{MSTseg1.jpg}
-\caption{MSTseg1.jpg}
-\end{figure}
-
+For this image, your function should return 495,012 edges.
+\label{problem4}
 \end{problem}
 
-You then build the minimum spanning tree.
-If you take away the edge with the greatest weight you will get two forests (removing an edge from any graph with no cycles will create two forests).
-Then you can continue doing this to each forest to split up those forests.
+Once we have converted our image into a graph, we can build the minimum spanning tree.
+If we take away the edge with the greatest weight from the tree we will get two \emph{forests} (removing an edge from any graph with no cycles will create two forests).
 Each forest corresponds to a segment of the image.
+If we remove the edge with the greatest weight from each of those forests, we will create new forests.
+
+We can use Kruskal's algorithm to find these forests.
 Let $k$ be the number of divisions that is wanted and $n$ be the number of nodes.
-Kruskal's algorithm is performed until $n-(k+1)$ edges are added which is the same as taking out the $k$ edges with greatest weights.
+We run Kruskal's algorithm until $n-(k+1)$ edges are added, which is the same as taking out the $k$ edges with greatest weights.
+So instead of building the whole minimum spanning tree and then removing the edges with greatest weights, we just apply Kruskal's algorithm up to a certain point and stop, giving us the forests or segments we want.
 
-The provided Kruskal's algorithm takes as inputs the list of nodes, the list of edges and the number of divisions desired.
-The number of divisions often has to be higher than the actual number that is needed because sometimes one or two pixels form a division because the difference between them and the pixels around them is so great.
-You will have to adjust the number of divisions until the desired result is found.  See Figure 5.6.
+Below is provided a modified version of the Kruskal's algorithm we did previously to find the minimum spanning tree of a graph.
+It takes as inputs the list of nodes and the list of edges from a graph, along with the number of divisions desired.
+The number of divisions often has to be set higher than the actual number that is needed because sometimes one or two pixels form a spurious division (the difference between them and the pixels around them is too great).
+You will have to adjust the number of divisions until the desired result is found.  See Figure \ref{fig:graph6}.
+
+\lstinputlisting[style=fromfile]{kruskalDiv.py}
 
 \begin{problem}
-Perform the image segmentation algorithm on the image, then graph the original image and the three largest divisions.
-(Use the Counter class from collections to find the number of pixels in each division.)
-\end{problem}
-
-\begin{problem}
-Make a division of the image a different color.
+Perform the image segmentation algorithm on the black and white image `MSTseg1.jpg'.
+Plot the image and the three largest segments using the plot function given below.
+This function takes in two inputs.
+The first is the values of the dictionary returned by the modified Kruskal's function given above.
+Make sure that you pass this in as a numpy array.
+The second input is the image array we are segmenting.
+\lstinputlisting[style=fromfile]{plotSeg.py}
 \end{problem}
@@
 \caption{The original image is in the top left hand corner. The three larges segments are shown in the other corners. The original image was 498x498 and 50000 divisions were used.}
-\label{mst:graph6}
+\label{fig:graph6}
 \end{figure}

### 2014-08-09 Algorithms/MST/plotSeg.py
Added a function that will help plot the image segmentations for Problem 5.

@@ -1,0 +1,15 @@
+def plotSeg(segments,img):
+    s=img.shape[0]
+    d=Counter(segments)
+    segs=d.most_common(10)
+    parts=segments.reshape((s,s))
+    plt.subplot(221)
+    plt.imshow(img)
+    plt.subplot(222)
+    plt.imshow(img*(parts==segs[0][0]))
+    plt.subplot(223)
+    plt.imshow(img*(parts==segs[1][0]))
+    plt.subplot(224)
+    plt.imshow(img*(parts==segs[2][0]))
+    plt.gray()
+    plt.show()

### 2014-08-20 Applications/ImageSegment/ImageSegment.tex
some grammer fixes
Added more directions and clarification

@@
 Figure \ref{fig:example_graph} shows a simple undirected graph.
-The graph in Figure \ref{fig:example_graph} can be represented by the following adjacency matrix.
+It can be represented by the following adjacency matrix.
 \[
@@
 \]
-For undirected graphs, the adjacency matrix will always be symmetric.
+The adjacency matrix will always be symmetric for undirected graphs.
 The diagonal represents self-edges, or edges that connect a node to itself.
@@
 For a directed graph, each node has an \emph{out-degree} (the number of edges directed away from a node) and an \emph{in-degree} (the number edges directed toward a node).
-The degree matrix of a graph is a diagonal matrix that has the degree of each node as the diagonal entries (thus, for a directed graph there is an in-degree matrix and an out-degree matrix).
+The \emph{degree matrix} of a graph is a diagonal matrix that has the degree of each node as the diagonal entries (thus, for a directed graph there is an in-degree matrix and an out-degree matrix).
 
-Another useful way of representing an undirected graph is a Laplacian matrix.
-This special matrix can reveal a lot of information about a graph.
+Another useful way of representing an undirected graph is a \emph{Laplacian matrix}.
+This special matrix can reveal a lot of information about a graph, which we will see later in this lab.
 \begin{definition}
@@
 \begin{problem}
-Calculate the Laplacian matrix of the graph in Figure \ref{fig:example_graph} by calculating $D$ and $A$.
-Think about how to find the number of neighboring nodes using an adjacency matrix.
+Write a function that accepts the adjacency matrix of a graph as an argument.
+Use the adjacency matrix to check that the graph is undirected and has no self-edges.
+Then using the definition above, calcluate the Laplacian matrix.
 
-Write a general solution that will calculate the Laplacian of any small graph.
-Use the adjacency matrix to check that the graph is unweighted, undirected, and contains no self-edges.
+Test your function by finding the Laplacian matrix of the graph in Figure \ref{fig:example_graph}. Your function should return the following Laplacian matrix:
+\[
+\begin{pmatrix}
+2 &-1 & 0 & 0 &-1 & 0 \\
+-1 & 3 &-1 & 0 &-1 & 0 \\
+0 &-1 & 2 &-1 & 0 & 0 \\
+0 & 0 &-1 & 3 &-1 &-1 \\
+-1 &-1 & 0 &-1 & 3 & 0 \\
+0 & 0 & 0 &-1 & 0 & 1 \\
+\end{pmatrix}
+\]
 \label{prob:laplacian}
@@
 
-We may also define weighted graphs so that each edge has a weight or cost attached to it.
-In this case the adjacency matrix will have the cost in each entry, rather than just a 1 or 0.
-An example of a weighted may be a collection of cities as vertices, roads connecting them as edges, and the distance of the connecting roads as the cost attached to each edge.
+We may also define \emph{weighted graphs} so that each edge has a weight or cost attached to it.
+In this case, the entries of the adjacency matrix are the costs, rather than just ones or zeros.
+An example of a weighted graph could be a collection of cities as vertices, roads connecting them as edges, and the distances of the connecting roads as the costs attached to each edge.
 
-In this lab we will be investigating the sorts of things that we can discover about a graph by examining the spectrum of its adjacency matrix.
-While the formulation of the adjacency matrix seems to be very simple, its structure allows us to learn surprising things by examining its eigenvalues.
+In this lab we will learn about the information we can discover about a graph from examining the spectrum of its adjacency matrix and its Laplacian.
+While the formulation of the Laplacian matrix seems to be very simple, its structure allows us to learn surprising things by examining its eigenvalues.
 
-For a weighted graph the definition of the Laplacian of a graph is:
+For a weighted graph the definition of the Laplacian is as follows.
 
@@
 The Laplacian matrix of a graph will typically be very sparse.
-We will be examining the spectrum of the Laplacian matrix.
 
-A connected graph is a graph where every vertex is connected to every other vertex by at least one path.
+A \emph{connected graph} is a graph where every vertex is connected to every other vertex by at least one path.
 An important question one might ask about a graph is whether or not it is connected.
@@
 A naive approach would be to exhaustively map every possible path from each vertex.
-While this would be feasible for very small graphs, most interesting graphs (for example, the internet) will have thousands of vertices and such an approach becomes essentially impossible to execute.
+While this would be feasible for very small graphs, most interesting graphs (for example, the internet) will have thousands of vertices, and such an approach becomes essentially impossible to execute.
 
@@
 \begin{problem}Write a function \li{laplacian} that accepts an adjacency matrix as an argument and returns the Laplacian matrix and its second smallest eigenvalue.
-Use the \li{scipy.linalg} package to compute this eigenvalue.
-Using {\tt numpy.rand(n,n)}, generate several random matrices of various sizes.
-Use masking to change the sparsity of said matrices, i.e. {\tt numpy.rand(n,n) > c} for {\tt c = 0.25,0.5,0.75}.
-Try different distributions and $c$ values.
-What can we expect about the connectivity of random matrices?
+Use the \li{scipy.linalg} package to compute the eigenvalue.
+
+Use \li{numpy.random.rand(n,n)} to generate two random matrices of sizes n=10 and n=100.
+Then use masking to change the sparsity of each of the matrices, i.e.
+\li{numpy.random.rand(n,n) > c} for \li{c = 0.25,0.95}.
+This will create four different matrices of boolean values.
+Remember that when doing calculations with boolean values that \li{True} is considered to be a 1 and \li{False} is considered to be a 0.
+Is the n=10 adjacency matrix connected when we let c=.25?
+What about when we let c=.95?
+Is the n=100 adjacency matrix connected when c=.95?
 \end{problem}
 
-A related problem is found in image processing.
+\section*{Image Segmentation}
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{monuments.png}
+\caption{An image and its segments.}
+\label{fig:monument}
+\end{figure}
+
+An important problem in image processing is that of image segmentation.
+When humans observe an image, they can easily pick out portions of an image that ``belong together.''
+For example, if we saw the picture of a person against a black background, the pixels making up the background would make up one segment of an image, and the person would make up the other part.
+While this seems simple for a human, how can we program a computer to do so? (See Figure \ref{fig:monument} for an example of a segmented image).
+
+First we need to know how to find the adjacency matrix of an image.
 An image is a collection of coordinates and light intensities.
@@
 	\item $\sigma_I$ and $\sigma_d$ are constants
+	\item $r$ is the radius
 \end{itemize}
 Thus, given a $N\times N$ image, we will produce an adjacency matrix of size $N^2\times N^2$.
-Even for smaller images, this will become very large, but we have sparsity on our side.
+Even for smaller images, this will become very large, but we have sparsity on our side, especially if $r$ is small.
 Even so, only do this for images smaller than 50 by 50.
+
+In the following problem we will write a function that finds the adjacency matrix of an image, based on the definition from \eqref{eq:adjacency}.
+To do this, we first flatten our image being passed in.
+
+\begin{lstlisting}
+nodes = img.flatten()
+\end{lstlisting}
+
+This allows us to index all of our pixels or nodes using an one-dimensional array.
+By doing this, we are better able calculate the weights in-between nodes.
+It is important to note here that this is only one way of indexing the nodes in our image.
+We can still index our nodes by using the tuple notation of matrix entries.
+So for example, given a $2 \times 2$ matrix we can access the bottom-right element, if we are using indexing in Python, by the tuple (1,1).
+If we choose to flatten our matrix we can simple use the index 3.
+
 Since each pixel is only connected to the pixels within radius $r$ around it, relatively few entries will be non-zero.
+For this reason, we will want to work with sparse matrices from the \li{scipiy.sparse} package.
+We initialize our adjacency matrix as a sparse matrix as follows:
+
+\begin{lstlisting}
+W = spar.lil_matrix((nodes.size, nodes.size), dtype=float)
+\end{lstlisting}
+
+We use \li{lil_matrix} because it gives us an efficent way to build our matrix piece by piece.
+To learn more about sparse matrices in scipy read the documenation at \url{http://docs.scipy.org/doc/scipy/reference/sparse.html}.
+
+Now we go through each node of our image matrix.
+A helper function has been provided that finds the pixels and corresponding distances within a given radius of a specific pixel.
+Note that the pixel or node being passed in is indexed in tuple form, as a given row and column number, while it returns an array of pixels with indices in a flat array form (a single number).
 
@@
 \end{problem}
-
-An important problem in image processing is that of image segmentation.
-When humans observe an image, they can easily pick out portions of an image that ``belong together.''
-For example, if we saw the picture of a person against a black background, the pixels making up the background would make up one segment of an image, and the person would make up the other part.
-While this seems simple for a human, how can we program a computer to do so?
 

### 2014-08-20 Applications/ImageSegment/ImageSegment.tex
Problem 4 I clarifies and provided helper code

@@
 When humans observe an image, they can easily pick out portions of an image that ``belong together.''
-For example, if we saw the picture of a person against a black background, the pixels making up the background would make up one segment of an image, and the person would make up the other part.
+For example, if we saw a picture of a person against a black background, the pixels making up the background would make up one segment of the image, and the person would make up the other part.
 While this seems simple for a human, how can we program a computer to do so? (See Figure \ref{fig:monument} for an example of a segmented image).
@@
 \end{equation}
-for $i = 1 \hdots N^2$ and $j = 1 \hdots N^2$ (traveling through each pixel in the image), where
+for $i = 1 \hdots N^2$ and $j = 1 \hdots N^2$ (where both $i$ and $j$ travel through each pixel in the image), and where
 \begin{itemize}
@@
 In the following problem we will write a function that finds the adjacency matrix of an image, based on the definition from \eqref{eq:adjacency}.
-To do this, we first flatten our image being passed in.
+To do this, we first flatten the image we are working on.
 
@@
 
-This allows us to index all of our pixels or nodes using an one-dimensional array.
-By doing this, we are better able calculate the weights in-between nodes.
-It is important to note here that this is only one way of indexing the nodes in our image.
-We can still index our nodes by using the tuple notation of matrix entries.
-So for example, given a $2 \times 2$ matrix we can access the bottom-right element, if we are using indexing in Python, by the tuple (1,1).
-If we choose to flatten our matrix we can simple use the index 3.
+This allows us to index all of the pixels (or nodes) using a one-dimensional array.
+By doing this, we are better able to calculate the weights in-between pixels.
+It is important to note here that this is only one way of indexing the pixels in our image.
+We can still index the pixels in our image or matrix by using standard tuple notation.
+So for example, given a $2 \times 2$ matrix we can access the bottom-right element (if we are including row 0 and column 0) by the tuple (1,1).
+If we choose to flatten our matrix we can index the same element by the number 3 (if we once again start at 0).
 
-Since each pixel is only connected to the pixels within radius $r$ around it, relatively few entries will be non-zero.
-For this reason, we will want to work with sparse matrices from the \li{scipiy.sparse} package.
-We initialize our adjacency matrix as a sparse matrix as follows:
+Since each pixel is only connected to the pixels within radius $r$, relatively few entries will be non-zero.
+For this reason, we will work with sparse matrices from the \li{scipiy.sparse} package.
+We first initialize our adjacency matrix as a sparse matrix.
 
@@
 \end{lstlisting}
+At this point we can initialize the main diagonal of the degree matrix.
+\begin{lstlisting}
+D = np.zeros((1, nodes.size))
+\end{lstlisting}
 
 We use \li{lil_matrix} because it gives us an efficent way to build our matrix piece by piece.
-To learn more about sparse matrices in scipy read the documenation at \url{http://docs.scipy.org/doc/scipy/reference/sparse.html}.
+To learn more about sparse matrices in scipy read the documentation at \url{http://docs.scipy.org/doc/scipy/reference/sparse.html}.
 
-Now we go through each node of our image matrix.
-A helper function has been provided that finds the pixels and corresponding distances within a given radius of a specific pixel.
-Note that the pixel or node being passed in is indexed in tuple form, as a given row and column number, while it returns an array of pixels with indices in a flat array form (a single number).
+Now we go through each pixel of our image matrix to calculate the weights.
+
+\begin{lstlisting}
+# 'height' is the original height of the image
+# 'width' is the original width of the image
+for row in xrange(height):
+    for col in xrange(width):
+        # Calculate the index of the pixel at (row, col) relative to the
+        # flattened array 'nodes'. 
+        rowcol = row * width + col
+\end{lstlisting}
+
+Say we start with pixel \li{A}.
+Notice that we do not need to find the weights between every pixel and pixel \li{A}, since most of the weights will be 0.
+We only need to find the weights between pixel \li{A} and those pixels that are within radius $r$ of pixel \li{A}.
+
+A helper function has been provided that finds the pixels and corresponding distances within radius $r$ of a single pixel.
+Note that the pixel being passed into the function is indexed as a tuple, or more precisely as a given row and column number.
+The function returns however, an array of pixels with indices in flat-array form (single numbers).
+This array contains the pixels that are within radius $r$ of the pixel passed into the function.
+\lstinputlisting[style=fromfile]{getNeighbors.py}
+Below we use the helper function to get the indices and distances.
+
+\begin{lstlisting}
+# find the indices and distancess of the pixels that are within 
+# distance r of the current pixel by calling getNeighbors
+indices, distances = getNeighbors(row, col, radius, height, width)
+\end{lstlisting}
+We can then calcluate the weights between pixels. Once we find the weights we can then add them into our adjacency matrix. 
+\begin{lstlisting}
+# 'weights' is the weights between the pixel we are on and the pixels within
+# radius r
+W[rowcol, indices] = weights
+\end{lstlisting}
+We can also calculate the main diagonal of the degree matrix.
+\begin{lstlisting}
+D[0,rowcol] = weights.sum()
+\end{lstlisting}
+
+After we have found our adjacency matrix (\li{W} in the code we have provided above) we convert it to a \li{csc} format. 
+\begin{lstlisting}
+W = W.tocsc()
+\end{lstlisting}
+The sparse matrix format \li{csc} is better for computations which we will want to use later.
 
@@
 $\sigma_I^2$ and $\sigma_d^2$, and returns the adjacency matrix defined in \eqref{eq:adjacency}.
-Notice that for each pixel you can save time by only checking the pixels $r$ rows and columns away.
+Test your function on the image \li{dream.png}.
+It is read in as a color photo so first convert it to grayscale to get rid of one of the dimensions.
+\begin{lstlisting}
+img_color = plt.imread('dream.png')
+img = (img_color[:,:,0]+img_color[:,:,1]+img_color[:,:,2])/3.0
+\end{lstlisting}
+Let $r=5.0$, $\sigma_I^2=.02$ and $\sigma_d^2=3.0$.
+
+%Notice that for each pixel you can save time by only checking the pixels $r$ rows and columns away.
 %For that you'll have to handle the pixels on the edges and corners of the image carefully.
 %I gave them new helper code, which abstracts away the edge cases.
-Make use of sparsity in order to feasibly store such a large matrix.
-We have provided code to help out with this problem.
 \end{problem}

### 2014-08-20 Applications/ImageSegment/getNeighbors.py
Helper function given in the lab

@@ -1,0 +1,24 @@
+def getNeighbors(row, col, radius, height, width):
+    '''
+    Calculate the indices and corresponding distances of pixels within radius
+    of the pixel at (row,col), where the pixels are in a (height, width) shaped
+    array. The returned indices are with respect to the flattened version of the
+    array. This is a helper function for adjacency.
+    Inputs:
+        row, col -- denotes the row and column number of the pixel we are 
+                    centered at.
+        radius -- radius of the circular region centered at pixel (row, col)
+        height, width -- the height and width of the original image, in pixels
+    Returns:
+        indices -- a flat array of indices of pixels that are within distance r
+                   of the pixel at (row, col)
+        distances -- a flat array giving the respective distances from these 
+                     pixels to the center pixel.
+    '''
+    r = int(radius)
+    x = np.arange(max(col - r, 0), min(col + r+1, width))
+    y = np.arange(max(row - r, 0), min(row + r+1, height))
+    X, Y = np.meshgrid(x, y)
+    R = np.sqrt(((X-np.float(col))**2+(Y-np.float(row))**2))
+    mask = (R<radius)
+    return (X[mask] + Y[mask]*width, R[mask])

### 2014-08-21 Applications/ImageSegment/ImageSegment.tex
Added a bunch of explanation for problem 5 and code. Also added a picture of a directed graph

@@
 \begin{figure}[h]
-\includegraphics[width=\textwidth/2]{graphExample}
+\includegraphics[width=\textwidth/3]{graphExample}
 \caption{A simple undirected graph}
 \label{fig:example_graph}
+\end{figure}
+
+\begin{figure}[h]
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.5cm,
+ thick,main node/.style={circle,draw}]
+
+  \node[main node] (A) [] {A};
+  \node[main node] (B) [below of=A] {B};
+  \node[main node] (C) [right of=A] {C};
+  \node[main node] (D) [below of=C] {D};
+  \node[main node] (E) [right of=C] {E};
+  \node[main node] (F) [right of=D] {F};
+
+  \foreach \s/\t in {A/C, B/A, B/D, C/E, C/F, D/C} {
+   \path[draw] (\s) edge (\t);}
+\end{tikzpicture}
+\caption{A simple directed graph}
 \end{figure}
@@
 An important problem in image processing is that of image segmentation.
-When humans observe an image, they can easily pick out portions of an image that ``belong together.''
+When humans observe an image, they can easily pick out portions of that image that ``belong together''.
 For example, if we saw a picture of a person against a black background, the pixels making up the background would make up one segment of the image, and the person would make up the other part.
@@
 
-First we need to know how to find the adjacency matrix of an image.
+One way to do this was presented in a paper by Jianbo Shi and Jitendra Malik in 2000. To use the algorithm they presented, we first need to find the adjacency matrix of an image.
 An image is a collection of coordinates and light intensities.
@@
 Thus, given a $N\times N$ image, we will produce an adjacency matrix of size $N^2\times N^2$.
-Even for smaller images, this will become very large, but we have sparsity on our side, especially if $r$ is small.
-Even so, only do this for images smaller than 50 by 50.
+Even for smaller images, this will become very large, but we have sparsity on our side, especially if $r$ is small. Even so, only do this for images smaller than 50 by 50.
+
+In definition \eqref{eq:adjacency}, we can think of $w_{ij}$ as the weight representing the similarity between pixels.
+So if a weight between two pixels is low, we can think of those pixels as being less similar and therefore more likely to represent where the image is segmented.
+We therefore want to segment our image along the edges with low weights.
+The total weight of the edges we remove is called the `cut' and so part of the mathematics behind image segmentation is minimizing the `cut', or in other words finding the segments that are least similar to each other.
 
@@
 We can still index the pixels in our image or matrix by using standard tuple notation.
-So for example, given a $2 \times 2$ matrix we can access the bottom-right element (if we are including row 0 and column 0) by the tuple (1,1).
+So for example, given a $2 \times 2$ matrix we can access the bottom-right element (if we are starting at row 0 and column 0) by the tuple (1,1).
 If we choose to flatten our matrix we can index the same element by the number 3 (if we once again start at 0).
 
-Since each pixel is only connected to the pixels within radius $r$, relatively few entries will be non-zero.
+Since each pixel is only connected to the pixels within radius $r$, relatively few entries of our adjacency matrix will be non-zero.
 For this reason, we will work with sparse matrices from the \li{scipiy.sparse} package.
@@
 \end{lstlisting}
-At this point we can initialize the main diagonal of the degree matrix.
+We use \li{lil_matrix} because it gives us an efficent way to build our matrix piece by piece.
+To learn more about sparse matrices in scipy read the documentation at \url{http://docs.scipy.org/doc/scipy/reference/sparse.html}.
+
+At this point we can initialize the main diagonal of the degree matrix. We only care about the main diagonal because the degree matrix is a diagonal matrix.
 \begin{lstlisting}
@@
 \end{lstlisting}
-
-We use \li{lil_matrix} because it gives us an efficent way to build our matrix piece by piece.
-To learn more about sparse matrices in scipy read the documentation at \url{http://docs.scipy.org/doc/scipy/reference/sparse.html}.
-
 Now we go through each pixel of our image matrix to calculate the weights.
@@
 \end{lstlisting}
-
 Say we start with pixel \li{A}.
@@
 A helper function has been provided that finds the pixels and corresponding distances within radius $r$ of a single pixel.
-Note that the pixel being passed into the function is indexed as a tuple, or more precisely as a given row and column number.
-The function returns however, an array of pixels with indices in flat-array form (single numbers).
-This array contains the pixels that are within radius $r$ of the pixel passed into the function.
+Note that the pixel being passed into the function is indexed as a tuple, or more precisely as a given row and column index.
+The function returns however, an array of indices in the flat array form (single numbers) of pixels within radius $r$ of the pixel passed into the function.
 \lstinputlisting[style=fromfile]{getNeighbors.py}
@@
 \begin{lstlisting}
-# find the indices and distancess of the pixels that are within 
+# find the indices and distances of the pixels that are within 
 # distance r of the current pixel by calling getNeighbors
@@
 \end{lstlisting}
-We can then calcluate the weights between pixels. Once we find the weights we can then add them into our adjacency matrix. 
+We can then calcluate the weights between pixels by using definition \eqref{eq:adjacency}.
+The helper function already calculates the distances, so now we just need to find the difference in brightness of pixels by subtracting the corresponding pixels.
+Once we find the weights we can then add them into our adjacency matrix. 
 \begin{lstlisting}
@@
 \end{lstlisting}
-We can also calculate the main diagonal of the degree matrix.
+We can also calculate part of the main diagonal of the degree matrix.
 \begin{lstlisting}
@@
 \end{lstlisting}
+We now loop to the next pixel and repeat the process.
 
-After we have found our adjacency matrix (\li{W} in the code we have provided above) we convert it to a \li{csc} format. 
+After we have found our adjacency matrix (\li{W} in the code we have provided above) we convert it to a \li{csc} format, and return both \li{W} and \li{D}. 
 \begin{lstlisting}
 W = W.tocsc()
+return W, D
 \end{lstlisting}
@@
 Test your function on the image \li{dream.png}.
-It is read in as a color photo so first convert it to grayscale to get rid of one of the dimensions.
+This image is read in as a color photo so first convert it to grayscale to reduce it to two dimensions.
 \begin{lstlisting}
+# read in the image
 img_color = plt.imread('dream.png')
+# convert to grayscale
 img = (img_color[:,:,0]+img_color[:,:,1]+img_color[:,:,2])/3.0
 \end{lstlisting}
-Let $r=5.0$, $\sigma_I^2=.02$ and $\sigma_d^2=3.0$.
+Let $r=5.0$, $\sigma_I^2=0.02$ and $\sigma_d^2=3.0$.
 
@@
 %I gave them new helper code, which abstracts away the edge cases.
+\label{prob4}
 \end{problem}
 
-Once again, we turn to the eigenvalues of the Laplacian matrix.
-We examine the second smallest eigenvalue of $D^{-\frac{1}{2}}QD^{-\frac{1}{2}}$, where $D$ is the degree matrix and $Q$ is the Laplacian matrix of the adjacency matrix defined in \eqref{eq:adjacency}.
-The associated $N^2 \times 1$ eigenvector will have positive and negative entries, splitting the image into two parts.
-Thanks to some very fancy mathematics and the way we have defined the weights between nodes, this does a pretty good job of breaking an image into two segments.
-Recursively executing this function, we can find segments within segments.
+As was mentioned before, we are trying to minimize the `cut', or the total weight of the edges we remove to create segments.
+This problem comes down to finding the second smallest eigenvalue of $D^{-1/2}QD^{-1/2}$, where $D$ is the degree matrix and $Q$ is the Laplacian matrix of the adjacency matrix defined in \eqref{eq:adjacency}.
+The mathematics behind this is quite involved, so the reader is encouraged to read the paper that Shi and Malik wrote at \url{http://www.cs.berkeley.edu/~malik/papers/SM-ncut.pdf} for a better understanding.
+
+To calculate $D^{-1/2}QD^{-1/2}$, we once again use sparse matrices.
+We use the \li{adjacency} function from Problem \ref{prob4} to find the adjacency matrix \li{W} and the main diagonal of the degree matrix \li{D}.
+We can convert \li{D} into a diagonal sparse matrix using \li{spar.spdiags}.
+
+\begin{lstlisting}
+Ds = spar.spdiags(D, 0, D.shape[1], D.shape[1], format = 'csc')
+\end{lstlisting}
+
+In like manner we can find the diagonal sparse matrix $D^{-1/2}$.
+We can also calculate the Laplacian matrix $Q$ since \li{Ds} and \li{W} are both sparse matrices and in `csc' format.
+To matrix multiply two sparse matrices together, use the \li{dot} function.
+For example, if \li{M} and \li{N} are two sparse matrices, use \li{M.dot(N)}.
+
+The eigenvector associated with the second smallest eigenvalue will be a $N^2 \times 1$ vector and will have positive and negative entries.
+If we reshape this vector to be the same size as our original image we can create a mask that will create two segments - one segment corresponding to the positive values of the eigenvector and the other segment corresponding to the negative values of the eigenvector.
+
+\begin{lstlisting}
+# create a mask array that is True wherever the eigenvector is positive.
+# reshape it to be the size of img.
+mask = (eigvec>0).reshape(img.shape)
+\end{lstlisting}
+To find the segments, we only need to take the product of our mask with the original image.
+
+\begin{lstlisting}
+# create the positive segment by masking out the pixels in img 
+# belonging to the negative segment.
+pos = img*mask
+\end{lstlisting}
+By recursively executing this function, we can find segments within segments.
+
 %Figure \ref{segmentation:example} shows a simple example of an image and its segments.
@@
 Accept an image array as an argument and return the two segments.
-Use $r = 5, \sigma_I^2 = 0.02,$ and $\sigma_d^2 = 3.0$.
-Remember that the Laplacian matrix will be very large but also very sparse.
-Because of the way we defined $\sigma_I^2$ the image matrix intensity values need to be between $0.0$ and $1.0$.
+Use $r = 5.0, \sigma_I^2 = 0.02,$ and $\sigma_d^2 = 3.0$.
+Remember that the Laplacian matrix will be very large, but also very sparse.
+Because of the way we defined $\sigma_I^2$ the image matrix intensity values need to be between $0.0$ and $1.0$. If this is not standard for the images you are using, change $\sigma_I^2$.
+
+Test your function on the image \li{dream.png}. Your segments should look like the segments below.
 \end{problem}

### 2014-08-21 Applications/ImageSegment/getNeighbors.py
Made a small change in the comments

@@ -1,3 +1,4 @@
+# helper function
 def getNeighbors(row, col, radius, height, width):
     '''
     Calculate the indices and corresponding distances of pixels within radius

### 2014-08-22 Python/StandardLibrary/stdlib.tex
Fixed the problem labels

\lab{Python}{The Standard Library}{Standard Library}
\objective{Become familiar with the Python standard library}
\label{lab:Standard Library}

Python has a very rich set of tools available by default.
There are about 80 built-in functions that are always present in any Python environment.
In addition to these 80 core functions, the standard library includes thousands of useful functions that cover almost every imaginable need.
Python has a ``batteries included'' philosophy that makes many complex tasks almost trivial to implement using the standard library.

The standard library is comprised of over a hundred different modules.
Python modules are simply Python source files where classes, functions, and variables are defined.
We can import any Python source file as a module and access the objects defined in it.
In this lab, we will look at the modules available in the Python standard library and at other built-in functions. The official documenation for the standard library can be found at \url{https://docs.python.org/2/library/}.

\section*{File Objects in Python}
One of the built-in functionalities of Python is working with files.
Python has a useful file object which acts as an interface to all kinds of different streams. Streams are simply sequences of data.
File objects are created using the built-in \li{open} command.

\begin{lstlisting}
f = open('filename.txt','r')
# This will iterate through your entire file and print each line.
for line in f:
        print line
f.close()
\end{lstlisting}

The \li{open} command accepts up to three arguments, or parameters: filename, mode, and buffering.
The mode determines the kind of access to use when opening the file.
Possible mode strings are:
\begin{description}
\item \li{'r'} Opens a file for read-only access.
This is the default mode.
\item \li{'w'} Opens a file for write-only access.
This mode creates the file if it doesn't already exist, and overwrites everything in the file if it does exist.
\item \li{'a'} Opens a file for appending.
The file pointer is at the end of the file, so any data already in the file will not be overwritten.
Creates a new file if none exists.
\end{description}
There are also two possible modifiers for the mode: \li{b} (binary) and \li{+} (read-write access).
Binary mode will open the file as-is, without any newline conversions.
The plus mode will open a file for reading and writing.
It is important to know that the modes \li{r+} and \li{w+} are not equivalent.
If your file cannot be opened for any reason, an exception is raised (usually an \li{IOException}).
Every file object has several attributes and methods (Table \ref{table:fileattribs}).
\begin{table}
\begin{tabular}{|l|l|}
\hline
Attribute & Description \\
\hline
\li{closed} & True if file object is closed. \\
\li{mode} & The access mode used to open the file object. \\
\li{name} & The name of the file. \\
\hline
\hline
Method & Description \\
\hline
\li{close()} & Flush any delayed writes and close the file object. \\
\li{flush()} & Flush any delayed writes.  File object remains open. \\
\li{read()} & Read the next string of the file. \\
\li{readline()} & Read a line of the file. \\
\li{readlines()} & Read lines of the file until end of file. \\
\li{seek(offset)} & Place the file pointer at a certain position within the file. \\
\li{tell()} & Return the current position of the file pointer. \\
\li{write()} & Write a string to the file. \\
\li{writelines()} & Write a sequence of strings to the file. \\
\hline
\end{tabular}
\caption{File object attributes and methods.}
\label{table:fileattribs}
\end{table}

Instead of using f = \li{open}, we usually open files using \li{with}.
Notice that when we use the \li{with} command we don't have to explicitly close the file;
the moment we exit the \li{with} block, the file is automatically closed safely.
We generally prefer to use the method below because it guarantees that our files will be closed properly, even if an unexpected error occurs.
\begin{lstlisting}
with open('filename.txt', 'r') as f:
    for line in f:
        print line
\end{lstlisting}

To write to a file, we do something similar to the following.
\begin{lstlisting}
num = 17
with open('output.txt', 'w') as f:
    f.write("My favorite number: ")
    # You can only write strings to a file, so you must convert non-string data to string format using the str() command before writing it to the file.
    f.write(str(num))
\end{lstlisting}

Note that we can declare variables inside the \li{with} block that are accessible outside that block.
\begin{lstlisting}
# Read your file using a with statement.
with open('filename.txt', 'r') as f:
        read_data = f.read()
# read_data is accessible outside of the with-block.
print read_data
\end{lstlisting}

\begin{problem}
Throughout this lab, we will write a script that simulates the Hunger Games tournament. \emph{The Hunger Games} is a popular novel about a dystopian society where each year boys and girls (called ``tributes") from different districts compete to the death in a tournament of survival. Attached with this lab is a text file named \li{events.txt} that has a list of events that might occur during a Hunger Games tournament. Write a function that reads in this file and returns a list of the events.

When you read in the file for the first time you may notice that the character `$\backslash$n' appears after every event. This happens because in the text file the events are on different lines, which the file object interprets as `$\backslash$n' after every event. To get rid of these characters use the .strip() method. For example:

\begin{lstlisting}
with open('filename.txt', 'r') as f:
    for line in f:
        print line.strip()
\end{lstlisting}

Note: Throughout this lab, keep your code in a .py file and run it from command line (use the command \li{python filename.py}), especially as you progress to later problems. You can still test solutions in an IPython notebook, but this will allow you to use all of the modules in this lab.

\end{problem}

\section*{Namespaces and Importing Modules}
Names are a fundamental concept in Python.
What we normally call \emph{variables} in other languages are \emph{names} in Python, but with a few key differences.
Names act as labels for Python objects in memory.
An object can have one or more names.
\begin{lstlisting}
a = 5
b = [1, 2, 3, 4]
\end{lstlisting}
In the code above, we define two Python objects in memory: an integer and a list.
We attach the name \li{a} to the integer object and the name \li{b} to the list object.
If we make the assignment \li{a = b}, the names \li{a} and \li{b} now both point to the same list object in memory.
Since functions in Python are simply objects, we can assign names for those too!
\begin{lstlisting}
def func():
    return 42
f = func
\end{lstlisting}
Whenever we want to call \li{func()}, we can call \li{f()} instead.

A \emph{namespace} is simply a collection of names.
A \emph{module} in Python is simply a Python file.
Each module gets its own namespace, so every name \emph{in that namespace} must be unique.
You can think of a module as simply a collection of names assigned to various Python objects.
Since every module has its own namespace, these namespaces are completely isolated; we can therefore have a function with the same name in two different modules without problems arising.
The main Python program gets a special namespace called \li{__main__}.
The name of a namespace is stored in the \li{__name__} attribute.
This is incredibly useful for checking to see if a namespace we are in has been imported, or is being run directly by the interpreter.
\begin{lstlisting}
if __name__ == "__main__":
    print "I am being run from the python interpreter."
elif __name__ != "__main__":
    print "I have been imported by another python module."
    print __name__
\end{lstlisting}

There are three ways to import names from a module.
\begin{description}
\item \li{import module [as alias]} This will import a module's namespace into the main namespace.
We can also use \li{as alias} to define an optional alias if the module name is longer than we would prefer to type repeatedly.
The method names in the module are accessed by \li{module.name}.
\begin{lstlisting}
# Import the module 'numpy', a scientific computing package.
# 'np' is the alias name.
import numpy as np
# 'linspace' is a method in the numpy module.
np.linspace(0, 10, 10)
\end{lstlisting}

\item \li{from module import name [as alias]} This will import a particular method name in the module directly into the main namespace.
This might replace a name in the main namespace if the imported name is not unique.
The names imported this way can be accessed directly by \li{name}.

\begin{lstlisting}
# Import the 'pyplot' method from the 'matplotlib' module
from matplotlib import pyplot as plt

\end{lstlisting}
\item \li{from module import *} This will import all names in the module's namespace directly into the main namespace.
This form is generally considered bad practice as it completely circumvents all the protections that namespaces provide, and can lead to many unknown name conflicts. Do not do this.
\end{description}

When importing modules it is good practice to put all of your import statements at the top of the script, and not scattered throughout or inside of functions.

\begin{warn}
Be careful when importing modules using the \li{from module import item} syntax.
Consider the following scenario.
Suppose we defined the following two functions in a file named \texttt{my\_functions.py}.
\begin{lstlisting}
def min(seq):
    return 0
def max(seq):
    return 1
\end{lstlisting}

We import the functions into our main program as follows
\begin{lstlisting}
from my_functions import min, max

# Calculate the median value of a sequence.
def calc_median(seq):
    return (min(seq)+max(seq))/2.
\end{lstlisting}

Suddenly our main program doesn't work as expected anymore!  What happened?
The functions \li{min()} and \li{max()} are defined in the core Python language.
However, when we imported the functions from our module, we overwrote Python's built-in \li{min()} and \li{max()} functions.
Now, whenever we call \li{min()}, it will always return 0!
In this case, there is no way to revert back to the Python functions without restarting the interpreter (in IPython, you restart the kernel).
This is a classic example of name collisions.
\end{warn}

\section*{\texttt{sys} Module}
In Python we will sometimes want to write standalone scripts that are executed from command line instead of the Python interpreter. The \li{sys} module helps with this by allowing us to access information specific to the system running Python.

For example, we often import \li{sys} to get the list \li{sys.argv} which is a list of arguments passed to the current environment.
It can be useful to access these values when writing programs that are run from command line.
Many programs are written to execute differently based on the various arguments and options specified at execution.
Here is an example.
\lstinputlisting[style=fromfile]{square.py}

When we run the script from the command line with a single argument, the code will print the name of the program along with the square of the number passed as the first argument.
If we execute the following script from the command line as follows:
\begin{lstlisting}[style=ShellInput]
python square.py 4
\end{lstlisting}
We should get the following output.
\begin{lstlisting}[style=ShellOutput]
square.py
16.0
\end{lstlisting}

\begin{problem}
As part of the Hunger Games simulation, we will write the proceedings of each day to a file. The name of the file will be given as an argument in command line when the Python script simulating the tournament is called. When calling the script from command line the input should look something like this:

\begin{lstlisting}[style=ShellOutput]
python hunger.py output.txt
\end{lstlisting}

\li{hunger.py} is the name of the script you are writing and \li{output.txt} is the name of the file (not yet created) where we will write the output of the tournament. In this problem, print the name of the file passed in (output.txt) to make sure the script recognizes the argument. Don't forget to import the \li{sys} module.
\label{prob:prob2}
\end{problem}

\section*{\texttt{csv} Module}
We will now look at a very useful module from the standard library for reading and writing data as comma separated values. We commonly use comma separated value (CSV) files to exchange data between databases and tables.
The \li{csv} module provides \li{reader} and \li{writer} objects. \li{DictReader} and \li{DictWriter} are analagous objects that use dictionaries to handle data.

Using these reader and writer objects, we can define the format of our CSV file.
Contrary to what the name implies, CSV files can be delimited with any character.
A delimiter is the special character that separates fields in a line. Commas are typical, but tabs and spaces are two other popular delimiting characters.
\begin{lstlisting}
# Set the delimiter to tabs.
csv_reader = csv.reader(csv_file, delimiter='\t')
\end{lstlisting}

We can also define the characters that separate fields, terminate lines, escape special characters, and enclose strings (\li{quotechar}.
The \li{csv} module refers to these special settings as \emph{dialects}.
The module comes with two dialects ready to use, excel (Windows Excel formatting) and excel-tab (Excel formatting, but tab-delimited).
\begin{lstlisting}
# Change the formatting dialect to excel.
csv_reader = csv.reader(csv_file, dialect='excel')
# Modify the formatting characters.
csv_reader = csv.reader(csv_file, quotechar='*', lineterminator='\n')

\end{lstlisting}

To print the contents of a CSV file, \li{test.csv}, we do the following.  A CSV reader will parse each line of a CSV file into a list of items.
\begin{lstlisting}
import csv

# Open test.csv as read-only.
with open('test.csv', 'r') as csv_file:
    # Create a csv reader object.
    csv_reader = csv.reader(csv_file)
    # Work with the reader object alone.
    for line in csv_reader:
        print line
\end{lstlisting}

Writing with a CSV \li{writer} object is very similar to writing with a regular file object.  Notice that we only need to pass a sequence of items to the CSV writer.  The CSV writer will take care of formatting things properly before writing it to the CSV file.
The following code demonstrates how this may be done.
\begin{lstlisting}
contents = [["Column 1", "Column 2", "Column 3"],
            [0,1,2], [3, 2, 1], [4,5,2], [68, 38, 99]]

# Open test_out.csv as a write-only file.
# Be careful! This will overwrite test_out.csv if it already exists.
with open('test_out.csv', 'w') as csv_file:
    # Create a csv writer object.
    csv_writer = csv.writer(csv_file)
    # Here, record is a loop variable that represents the rows.
    for record in contents:
        # Write rows using the writer object.
        csv_writer.writerow(record)
\end{lstlisting}

\begin{problem}
Attached with this lab is a CSV file that contains two columns. The first column lists the male tributes and the second column lists the female tributes for the Hunger Games. Using the CSV module, write a function that reads in both columns and seperates the male and female tributes into two lists. Return the lists. Don't forget to import the CSV module.

\emph{Hint:} We can access the different columns using a for loop with the following code:

\begin{lstlisting}
for male, female in csv_reader:
\end{lstlisting}

\end{problem}

\section*{\texttt{math} Module}
The \li{math} module and its companion, \li{cmath} (for complex numbers), are very valuable modules.
Common mathematical functions are defined in these modules, including \li{cos}, \li{sin}, \li{log}, and \li{sqrt}.

\begin{lstlisting}
import math
# Return the sine of 4 radians.
math.sin(4)
\end{lstlisting}

To learn more about the cmath module, see the official Python documentation at \url{https://docs.python.org/2/library/cmath.html}.

\begin{problem}
Import the \li{math} module and \li{cmath} module. Write a function that prints out both the complex and floating point results for $\sqrt{5}$.  What are the results for $\sqrt{-5}$?
\end{problem}

\section*{\texttt{random} Module}
The \li{random} module contains many helpful functions for obtaining random numbers.
These ``random'' numbers are not actually completely random; however, they are seemingly-random enough for most applications.
Some of the more commonly used methods of the \li{random} module are shown below.
\begin{lstlisting}
>>> import random

# Get a random integer from the interval [5, 12].
>>> random.randint(5, 12)

# Get a random integer from a range of integers [5, 12).
>>> random.randrange(5, 12)

#Get a floating point number between 0.0 and 1.0
random.random()

# Get a random element from a sequence. 
# In this example, random.choice will return a character from the string given.
>>> random.choice("This is a sentence")

# Get a random sample of length n from a sequence.
>>> random.sample(range(50), 10)
>>> [21, 25, 9, 30, 22, 41, 12, 26, 47, 5]

# Randomly shuffle a sequence in place.
>>> R = range(15)
>>> random.shuffle(R)
>>> print R
[7, 9, 13, 3, 5, 8, 10, 2, 0, 6, 4, 11, 1, 12, 14]

\end{lstlisting}

\begin{info}
The \li{random} module has functions that can sample from a variety of different statistical distributions including uniform, normal, beta, gamma, exponential, and so on. To learn more about random numbers in Python see the documentation at \url{https://docs.python.org/2/library/random.html}.
\end{info}

\begin{problem}
Each contestant of the Hunger Games must survive each day if they are to win. Write a function that creates a list of 24 random floating point numbers between 1.0 and 10.0. The numbers in this list will be used later as the likelihood of the tributes surviving an event.
\label{prob:prob5}
\end{problem}

\section*{\texttt{itertools} Module}
Itertools is a very powerful module in the Python standard library. It is built around the concept
of generators and iterators.  The functions in \li{itertools} are fast and memory efficient and are designed to be used as
building blocks in larger functions. 
This efficiency is especially important in Python, where code is generally executed more slowly than in a compiled language like C or Fortran.

\subsection*{Generators}
Before discussing this module, you need to understand what a generator is, and the difference between generators and iterators.

Lists, tuples, sets and dictionaries are all examples of sequences in Python. It is frequently useful and necessary to visit all the elements of a sequence once; the process of doing so is called \emph{iteration}.
Each of the Python types we have mentioned define their own iterators, which you use every time you execute the statement \li{for <elem> in <sequence>}.

Python also has a special type of iterator object called a \emph{generator}.
Like a standard iterator, a generator returns a sequence of values, but unlike a standard iterator, which computes an entire sequence and returns values from it, a generator computes and returns the next value in the sequence each time it is called and discards the previous values. It never has to store all the values in a sequence. When is this helpful?

\begin{itemize}
\item \emph{Iterating through only part of a sequence.}
It is inefficient to create an entire sequence if we know that we will not need all of it.
Representing the sequence as a generator can avoid excess memory use and computation.
\item \emph{Iterating through a sequence once.} Consider the statement
\li{sum([i for i in range(1000) if i\%2 == 0])}.
We are creating two sequences just to iterate through each them once and never use them again.
We can make this more efficient using generators:
\li{sum(i for i in xrange(1000) if i\%2 == 0)}
Notice that we have used syntax similar to a list comprehension.
The line \li{(i for i in xrange(1000) if i\%2 == 0)} will define a generator object similar to the list made by \li{[i for i in xrange(1000) if i\%2 == 0]}.
The solution using generators will often execute faster, and it will almost always be more memory efficient.  Consider using generators for any function that reduces a sequence to a single value.  Examples of such functions are \li{min()}, \li{max()}, and \li{sum()}.
\item \emph{Calculating large sequences.}  A sequence must be stored somewhere in computer memory.
If the sequence is large, we could exhaust all available memory.
\item \emph{Calculations involving infinite sequences.}  Pre-computed sequences are necessarily finite; we simply cannot do computations involving infinite sequences with normal iterators. We cannot create a list that stores all natural numbers, but we can create a generator that returns the next natural number every time it is called.
\end{itemize}

Now that we understand what generators are and why they are important, we can discuss how to create and use them. Python uses the \li{yield} keyword to define a generator. As an example, consider the following, where we re-implement Python's \li{range} function as an iterator and a generator.
\begin{lstlisting}
def range_iter(start, stop, step=1):
    i = start
    r = []
    while i < stop:
       r.append(i)
       i = i + step
    return r
\end{lstlisting}
\begin{lstlisting}
def range_gen(start, stop, step=1):
    i = start
    while i < stop:
        yield i
        i = i + step
\end{lstlisting}
These two functions look similar, but they behave very differently.
The first function when executed will build a list one element at a time until finished,
at which point it will return the entire list.  On the author's computer, this function takes about 1.43ms
to build and returns a list of 10000 elements.  If we build the entire list using the generator function then the result is only marginally better, returning the 10000 elements in approximately 1.09ms.

But if we iterate through the list compared to the generator, we see a huge difference the first time each is called. The first function requires 1.43ms to execute the first time.  The second function,
a mere .00000041ms!  The reason for this is that the first function calculates its results and returns them all at once. The second function only creates a \emph{generator} object.

Generator objects have several methods.
The most important and most useful method is the \li{next()} method.  Every time this method is called, 
the generator resumes from its previous state and computes the next value in the sequence. After computing and yielding this value, the generator is suspended until \li{next} is called again.

When you are writing \li{for} loops, you should use a generator whenever it is reasonable to do so.
The \li{xrange()} function that is built into Python 2.7 is a generator-based version of \li{range()} and is generally faster when used in \li{for} loops; you should make a habit of using \li{xrange()} instead of \li{range()}. In a for loop, you can use \li{xrange()} the same way you would use \li{range()}.

\begin{info}
In Python 3, range is a generator by default, and when you actually need a list you must use something like list(range(x)), where x is an integer.
\end{info}

In general, the generators in \li{itertools} are used like this:
\begin{lstlisting}
import itertools
a = itertools.count(5) 
# Returns natural numbers, starting at 5, increasing by 1 each time next() is called.
a.next()
\end{lstlisting}

Note that while \li{count()} takes an integer as input, most generators work over \emph{iterables}, or any Python object that can be iterated through; this includes strings, lists, tuples, and so on.

\subsection*{Useful Generators in \texttt{itertools}}
The \li{itertools} module contains three main types of generators: infinite generators, shortest sequence generators, and combinatoric generators.
Some common generators are summarized in Table \ref{table:populargens}. Note that \li{islice} is particularly useful for efficiently iterating through only part of a sequence. For the argument values of and further information about these generators, as well as recipes for other useful generators, see the documentation at \url{http://docs.python.org/2/library/itertools.html}
\begin{table}
\begin{tabular}{|l|l|}
\hline
Generator & Example \\
\hline
\li{count()} & Count to infinity \\
\li{cycle()} & Cycle through a sequence indefinitely\\
\li{repeat()} & Repeat the input element indefinitely (or up to $n$ times) \\
\hline
\li{chain()} & \li{chain('ABC', 'DEF') --> A B C D E F} \\
\li{compress()} & \li{compress('ABCDEF', [1,0,1,0,1,1]) --> A C E F} \\
\li{islice()} & \li{islice('ABCDEFG', 2, None) --> C D E F G} \\
\li{imap()} & \li{imap(pow, (2,3,10), (5,2,3)) --> 32 9 1000} \\
\li{izip()} & \li{izip('ABCD', 'xy') --> Ax By} \\
\hline
\li{product()} & Return the Cartesian product of two iterables\\
\li{permutations()} & Return permutations of a sequence\\
\li{combinations()} & Return combinations of a sequence\\
\hline
\end{tabular}

\caption{Popular generators in \li{itertools} module.}
\label{table:populargens}
\end{table}

\section*{\texttt{collections} Module}
This module defines several specialized data structures to use in addition to the built-in Python data structures.

Named tuples are designed to help improve code readability in some cases.
Standard tuples in Python are accessed by index while named tuples allow access via index or by fieldname (which defines the tuple elements and acts like a key).
\begin{lstlisting}
>>> from collections import namedtuple

# Initialize namedtuple with typename of 'Stats'
# Use namedtuple('typename', 'fieldname1, fieldname2, etc.')
>>> player = namedtuple('Stats', 'shots, assists')

>>> playerA = player(18,12)

>>> playerA
>>> Stats(shots=18, assists=12)

# Access 'shots' for playerA
>>> playerA.shots
18
\end{lstlisting}

\label{deques}
A double-ended queue, or deque (pronounced ``deck''), can be thought of as a deck of cards.
Inserting and removing elements from either end is highly efficient.
Python's deque implementation only allows insertions at the left and right ends of the data structure (which is standard for deques).
This differs from a list, which allows insertions anywhere, but is very inefficient for all but right end insertions.

\begin{lstlisting}
>>> from collections import deque

>>> d = deque(range(10))

>>> d.appendleft(-1)

>>> print d
deque([-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9])

>>> d.rotate(1)

>>> print d
deque([9, -1, 0, 1, 2, 3, 4, 5, 6, 7, 8])
\end{lstlisting}

Counter objects are very efficient at counting items.  They behave like a Python dictionary.  Counts are allowed to be any integer, including 0 and negative values.

\begin{lstlisting}
>>> from collections import Counter

>>> letter_count = Counter()

>>> for char in "Mississippi":
>>>     letter_count[char] += 1

>>> letter_count
Counter({'i': 4, 's': 4, 'p': 2, 'M': 1})
\end{lstlisting}

Ordered dictionaries are exactly like standard dictionaries except for one important difference:
ordered dictionaries remember the order in which key-value pairs were added to the data structure.
When iterating over an ordered dictionary, the items are returned in the order they were added from first to last.

\begin{lstlisting}
>>> from collections import OrderedDict

>>> d = OrderedDict()

>>> d['a'] = 1

>>> d['b'] = 2

>>> print d
OrderedDict([('a', 1), ('b', 2)])

\end{lstlisting}

For further information and examples of how to use these container objects, see section 8.3 of the official Python tutorial (\url{https://docs.python.org/2/library/collections.html}).

\begin{problem}
In order to organize the tributes competing in the Hunger Games tournament, create a named tuple that contains information about each of the tributes including their name, district and gender. Then create a list named ``tributes'' that will hold all of the named tuples.

To do this, write a function that pairs the male and female tributes by district. Use the \li{izip()} generator from the \li{itertools} module together with the built-in function \li{enumerate}. The \li{enumerate} function provides an easy way to assign numbers in a for loop.
Put all together, the for loop should look something like this:

\begin{lstlisting}
for d, t in enumerate(itertools.izip(males, females),1):
        tributes.append(Tribute(t[0], d, "M"))
        tributes.append(Tribute(t[1], d, "F"))
\end{lstlisting}

where ``Tribute" is the named tuple, ``d" is the current district number provided by enumerate (starting at 1), and t is the \li{izip()} generator containing the male, female pair. Return the tributes list. The list should contain 24 named tuples, each one representing a tribute.
\label{prob:prob6}
\end{problem}

\section*{\texttt{timeit} Module}
This module is used to time the execution of small bits of Python code.
It is helpful to time lines of code in Python using the \li{timeit} module because it avoids a number of common pitfalls in measuring execution time; for example, it limits errors caused by startup and shutdown times by running the code repeatedly. In IPython, we usually time code using the built-in \li{\%timeit} function. Prefacing a line of code with \li{\%timeit} times that line; to time an entire cell of code, preface it with \li{\%\%timeit}. However, this will not work in all environments, notably in a command line setting, so it is important to know how to time code without the \li{\%timeit} function.

We will now define a function that times the execution of another function \li{f} and returns the minimum runtime for one call of \li{f}. This will be useful in other labs when you are asked to time the execution of your solutions. Before we can do this, however, we need to know what a wrapper function is. A wrapper function wraps a function and its arguments into a function object that can be called by the \li{timeit} function. In Python, this is done by declaring a \li{lambda} function, which is a temporary, anonymous function.

\begin{lstlisting}
import timeit

# time_func takes as its arguments the name of a function f, a tuple of the
# arguments of f, a dictionary of the keyword arguments of f, and two 
# keyword arguments.
def time_func(f, args=(), kargs={}, repeat=3, number=100):
	# Wrap f into pfunc.
	pfunc = lambda: f(*args, **kargs)
	# Define an object T that times pfunc once.
	T = timeit.Timer(pfunc)

	# Time f several times, return the name of f and the minimum runtime.
	try:
		# Repeat is also a timeit module function
		t = T.repeat(repeat=repeat, number=int(number))
		runtime = min(t)/float(number)
		return runtime
	# Print an error statement if something goes wrong.
	except:
		T.print_exc()
\end{lstlisting}

In the above code you may have noticed a \li{try} and \li{except} block. How this works is Python attempts to run the code following \li{try} and if that fails it skips to the \li{except} block and runs the code that is found there. In the above case if Python fails to run the code following \li{try}, it will jump to the \li{except} block and print an error message.

Remember that keyword arguments are arguments that are given a default value in the function declaration after the positional (non-keyword) arguments, but can be overwritten. Therefore, when calling \li{time_func}, the dictionary of keyword arguments of \li{f} (where the names of the keyword arguments are the keys) is optional, and so are the keyword arguments of \li{time_func} itself.
\li{repeat} defines the number of times to test the code, while \li{number} defines the number of times the code is run per test. If your code is small and simple, you can set the value of \li{number} very high without causing problems, but for bulky, slower code, \li{number} will need to be smaller, or you may have to wait a long time to get your timing results.

Here is an example of how you would use the above function in a script.

\begin{lstlisting}
# We will time how long it takes to run the following function.
def square(x):
    return x**2

# Call time_func where 4 is the parameter we are passing into 'square'.
# Notice that 4 is put in brackets.
time_func(square, [4])
\end{lstlisting}

For further information on the \li{timeit} module, see the documentation at \url{http://docs.python.org/2/library/timeit.html}.

\begin{problem}
Write two functions that will rotate the elements of a deque (from the collections module) and a list respectively.
\begin{enumerate}
\item Use the rotate method of a deque.
\item Use a for-loop to rotate a list.
\end{enumerate}

To rotate the list, remove elements from the right end one by one and insert them on the left end one by one.
Compare the timings you obtain from a deque and a list of 10000 elements, using the timer function defined in this section. Rotate all elements of the list and deque so that your final sequence is the same as when you started.
\end{problem}

\begin{problem}
Throughout this lab we have worked on simulating the Hunger Games tournament using the standard library and other built-in functions found in Python.

To piece everything together, write a function that takes in as parameters the list of events, the random list created in Problem \ref{prob:prob5}, and the list of tributes created in Problem \ref{prob:prob6}.

To simulate the game, you will need to do at least the following.
For each day of the simulation, each (surviving) tribute will experience one event per day. Each event should be randomly chosen from the list of events you read from the text file.
To determine the survival of a given tribute, randomly generate a real number between 1.0 and 10.0. 
If the random number is greater than the likelihood of surviving the event, then the tribute survives the event for that day. Note that the likelihood of surviving the event comes from the list of random numbers generated in Problem \ref{prob:prob5}. So the nth tribute/survivor will survive if their random number is greater than the nth element of the list of likelihoods from Problem \ref{prob:prob5}.

After each day, write the results of the competition to a file using the file object from the first part of this lab:

\begin{lstlisting}
with open('output.txt', 'w') as f:
\end{lstlisting}

The file \li{output.txt} is the name of the file you should have passed in from Problem \ref{prob:prob2} using the \li{sys} module. Write to the file the names of the survivors along with what events they survived, as well as the names of those who didn't survive and their associated events.

Make sure that you initialize your file object outside any loops, so that you don't completely overwrite your file each time you loop.

Stop the simulation when there is at most one surviving tribute and print the winner to standard output (it will appear in the console) and to your file.  It is possible that there are no surviving tributes.  If there are no surviving tributes, indicate so in your outputs.

Keep in mind good coding practice as you solve this problem. Use comments to document what your code is doing, and most importantly, plan out how you are going to approach the problem \emph{before} you actually start coding.

\end{problem}

\section*{Specifications}
We suggest that you submit your \li{solutions.py} file using the following format.
\lstinputlisting[style=fromfile]{solution_specs.py}
